---
title: "Getting Started with hurdlebb"
author: "JoonHo Lee"
date: "`r Sys.Date()`"
description: >
  A hands-on introduction to fitting Hurdle Beta-Binomial models with hurdlebb:
  explore data, specify a model, run MCMC diagnostics, posterior predictive checks,
  and cross-validation --- all using pre-computed results.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with hurdlebb}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
library(hurdlebb)
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width  = 7,
  fig.height = 5,
  fig.align  = "center",
  out.width  = "85%"
)
set.seed(42)

# Project colour palette
pal <- c(extensive = "#4393C3", intensive = "#D6604D",
         reference = "#1B7837", dispersion = "#762A83")

# Helper: find extdata files in installed package OR source tree
find_extdata <- function(filename) {
  path <- system.file("extdata", filename, package = "hurdlebb")
  if (nzchar(path) && file.exists(path)) return(path)
  path2 <- file.path("..", "inst", "extdata", filename)
  if (file.exists(path2)) return(path2)
  ""
}

# Guard: pre-computed results must exist for eval'd chunks
has_data <- nzchar(find_extdata("vig01_results.rds"))
```

## Overview

Many childcare datasets share a common structure: a large fraction of
providers do not serve a particular age group at all, while those that do
vary widely in how many children they enrol. Standard regression models
struggle with this combination of excess zeros and bounded, overdispersed
counts. The **hurdlebb** package provides a principled solution through the
Hurdle Beta-Binomial model --- a two-part framework that separates the
decision to participate from the intensity of participation.

In this vignette you will learn how to:

1. **Explore** a childcare enrolment dataset and recognize why a two-part
   model is needed.
2. **Specify** a Hurdle Beta-Binomial model using the `hbb()` formula
   interface.
3. **Interpret** the model summary, paying special attention to how the same
   covariate can push in opposite directions across the two margins.
4. **Diagnose** convergence with MCMC trace plots.
5. **Evaluate** model fit through posterior predictive checks (PPC).
6. **Compare** models using approximate leave-one-out cross-validation (LOO).
7. **Extract** fitted values and residuals for further diagnostics.

Because the MCMC sampler can take several minutes, this vignette uses
**pre-computed results** that ship with the package. Every fitting command is
shown so you can reproduce the analysis on your own machine, but the displayed
output comes from saved objects.


## The Data

The package includes `nsece_synth_small`, a synthetic dataset modelled after
the 2019 National Survey of Early Care and Education (NSECE). It contains
504 centre-based childcare providers drawn from 51 states (including DC).

```{r load-data}
data("nsece_synth_small", package = "hurdlebb")
str(nsece_synth_small)
```

Each row is a provider. The key columns are:

| Column     | Description                                            |
|:-----------|:-------------------------------------------------------|
| `y`        | Number of infant/toddler (IT) children enrolled        |
| `n_trial`  | Total capacity (children age 0--5)                     |
| `z`        | Binary: does the centre serve any IT children?         |
| `it_share` | Ratio `y / n_trial` (0 for non-servers)                |
| `poverty`  | Neighbourhood poverty rate                             |
| `urban`    | Urban location indicator                               |

Two features of this dataset motivate the Hurdle Beta-Binomial approach.
First, a substantial share of centres do not serve infants or toddlers at all:

```{r zero-rate}
cat("Structural zeros:", sum(nsece_synth_small$z == 0), "of",
    nrow(nsece_synth_small),
    sprintf("(%.0f%%)\n", 100 * mean(nsece_synth_small$z == 0)))
```

Second, among centres that *do* serve IT children, the enrolment share
varies considerably:

```{r it-share-summary}
servers <- nsece_synth_small[nsece_synth_small$z == 1, ]
summary(servers$it_share)
```

This combination --- a spike at zero plus a dispersed distribution over
$(0, 1]$ --- is the hallmark of a hurdle problem. The plots below make
the pattern vivid.

```{r fig-enrolment, fig.height = 4, fig.cap = "IT enrolment distribution. The spike at zero (blue) reflects centres that do not serve infants and toddlers. Positive counts (red) show the enrolment distribution among participating centres."}
if (requireNamespace("ggplot2", quietly = TRUE)) {
  library(ggplot2)
  dat <- nsece_synth_small
  dat$group <- ifelse(dat$y == 0, "Zero", "Positive")
  zero_pct <- sprintf("%.0f%% zeros", 100 * mean(dat$y == 0))

  ggplot(dat, aes(x = y, fill = group)) +
    geom_histogram(binwidth = 1, boundary = -0.5, colour = "white",
                   linewidth = 0.3) +
    scale_fill_manual(
      values = c(Zero = pal["extensive"], Positive = pal["intensive"]),
      guide  = "none"
    ) +
    annotate("text", x = max(dat$y) * 0.7, y = sum(dat$y == 0) * 0.85,
             label = zero_pct, size = 5, fontface = "bold",
             colour = pal["extensive"]) +
    labs(x = "IT enrolment count (y)", y = "Number of centres",
         title = "IT Enrolment Distribution") +
    theme_minimal(base_size = 12) +
    theme(panel.grid.minor = element_blank())
}
```

```{r fig-it-share, fig.height = 4, fig.cap = "IT share among serving centres. The Beta-like distribution motivates the Beta-Binomial intensive margin."}
if (requireNamespace("ggplot2", quietly = TRUE)) {
  mean_share <- mean(servers$it_share, na.rm = TRUE)

  ggplot(servers, aes(x = it_share)) +
    geom_histogram(aes(y = after_stat(density)), bins = 25,
                   fill = pal["intensive"], colour = "white",
                   linewidth = 0.3, alpha = 0.85) +
    geom_vline(xintercept = mean_share, linetype = "dashed",
               colour = pal["reference"], linewidth = 0.8) +
    annotate("text", x = mean_share + 0.04, y = Inf, vjust = 1.5,
             label = sprintf("mean = %.2f", mean_share),
             colour = pal["reference"], size = 4, fontface = "italic") +
    labs(x = "IT share (y / n)", y = "Density",
         title = "IT Share Among Serving Centres") +
    theme_minimal(base_size = 12) +
    theme(panel.grid.minor = element_blank())
}
```


## Specifying the Model

The Hurdle Beta-Binomial decomposes the outcome into two parts:

- **Part 1 (Extensive margin):** A Bernoulli model for whether the centre
  serves IT children at all. The participation probability $q_i$ is linked to
  covariates through a logit: $\text{logit}(q_i) = X_i'\alpha$.
- **Part 2 (Intensive margin):** A zero-truncated Beta-Binomial for the
  *number* of IT children enrolled, conditional on serving at least one. The
  expected IT share $\mu_i$ is also linked to covariates through a logit:
  $\text{logit}(\mu_i) = X_i'\beta$, with a dispersion parameter $\kappa$.

In **hurdlebb**, you specify both margins through a single formula. The
left-hand side uses the `y | trials(n_trial)` syntax to declare the count and
its upper bound:

```r
# This is the code you would run (takes ~2-4 minutes):
fit <- hbb(
  y | trials(n_trial) ~ poverty + urban,
  data   = nsece_synth_small,
  chains = 4,
  seed   = 42
)
```

The same set of predictors (`poverty` and `urban`) enters both the
extensive-margin equation (coefficients $\alpha$) and the intensive-margin
equation (coefficients $\beta$). This shared-covariate design makes it
straightforward to compare whether a variable pushes participation and
intensity in the same or opposite directions.


## Loading Pre-computed Results

For this vignette we load saved output instead of running the sampler:

```{r load-results, eval = has_data}
results <- readRDS(find_extdata("vig01_results.rds"))
```

The `print()` method gives a compact overview:

```{r show-print, eval = has_data}
cat(paste(results$print_text, collapse = "\n"))
```


## Model Summary

The `summary()` output presents posterior means, standard deviations, and 95%
credible intervals for every parameter:

```{r show-summary, eval = has_data}
cat(paste(results$summary_text, collapse = "\n"))
```

### Reading the coefficient table

The parameters fall into three groups:

- **Extensive-margin coefficients** ($\alpha_1, \alpha_2, \alpha_3$): Govern
  the log-odds of serving IT children. A positive $\alpha$ means the covariate
  *increases* the probability of participation.
- **Intensive-margin coefficients** ($\beta_1, \beta_2, \beta_3$): Govern the
  log-odds of IT *share* among servers. A positive $\beta$ means the covariate
  raises the expected fraction of capacity devoted to IT.
- **Dispersion** ($\log\kappa$): Controls extra-binomial variation. Larger
  $\kappa$ means less overdispersion.

You can extract coefficients programmatically:

```{r coef-all, eval = has_data}
results$coef_both
```

Or select a single margin:

```{r coef-margins, eval = has_data}
results$coef_ext   # extensive margin (alpha)
results$coef_int   # intensive margin (beta)
```

### The poverty reversal

The most striking pattern is the poverty covariate. The signs are *opposite*
across the two margins:

```{r poverty-reversal, eval = has_data}
cat(sprintf("Extensive margin (alpha_poverty): %+.3f\n",
            results$coef_both["alpha_poverty"]))
cat(sprintf("Intensive margin  (beta_poverty): %+.3f\n",
            results$coef_both["beta_poverty"]))
```

Higher neighbourhood poverty is associated with *lower* probability of serving
IT children ($\alpha < 0$), yet among centres that do serve, higher poverty is
associated with a *larger* IT share ($\beta > 0$). This **poverty reversal**
is the central substantive finding: poverty discourages entry into IT care but
intensifies enrolment among those that participate.

A standard single-equation model would blend these two opposing forces into one
ambiguous coefficient. The hurdle framework lets us see each margin clearly.


## Visualizing Coefficients

A forest plot displays all parameters with credible intervals, coloured by
margin. In a live session you would call `plot(fit, type = "coefficients")`.

```{r fig-forest, eval = has_data, fig.height = 4.5, fig.cap = "Coefficient estimates with 95% credible intervals. Blue = extensive margin, red = intensive margin, purple = dispersion. The dashed green line marks zero. Note the opposing signs for poverty across the two margins."}
if (requireNamespace("ggplot2", quietly = TRUE)) {
  library(ggplot2)

  fe <- results$summary_obj$fixed_effects
  fe$margin_group <- ifelse(grepl("^alpha", fe$parameter), "Extensive",
                    ifelse(grepl("^beta",  fe$parameter), "Intensive",
                           "Dispersion"))
  fe$margin_group <- factor(fe$margin_group,
                            levels = c("Extensive", "Intensive", "Dispersion"))
  fe$label <- fe$parameter
  fe$label <- factor(fe$label, levels = rev(fe$label))

  ggplot(fe, aes(x = estimate, y = label, colour = margin_group)) +
    geom_vline(xintercept = 0, linetype = "dashed",
               colour = pal["reference"], linewidth = 0.6) +
    geom_pointrange(aes(xmin = ci_lower, xmax = ci_upper),
                    size = 0.5, linewidth = 0.7, fatten = 2.5) +
    scale_colour_manual(
      values = c(Extensive  = pal["extensive"],
                 Intensive  = pal["intensive"],
                 Dispersion = pal["dispersion"]),
      name = "Margin"
    ) +
    labs(x = "Posterior estimate", y = NULL,
         title = "Coefficient Estimates with 95% Credible Intervals") +
    theme_minimal(base_size = 12) +
    theme(legend.position = "bottom",
          panel.grid.minor = element_blank())
}
```


## MCMC Diagnostics

Before trusting any Bayesian result, we need to verify that the sampler has
converged. The trace plot shows sampled values across iterations for each
chain. Well-mixed chains should look like "hairy caterpillars" --- dense,
overlapping bands with no trends or drifts.

```{r fig-trace, eval = has_data, fig.height = 5, fig.cap = "MCMC trace plots for four key parameters across four chains. Overlapping, stationary traces indicate reliable convergence."}
if (requireNamespace("ggplot2", quietly = TRUE)) {
  library(ggplot2)

  draws <- results$draws_df
  key_params <- c("alpha[1]", "alpha[2]", "beta[2]", "log_kappa")
  param_display <- c(
    "alpha[1]"  = "Intercept (extensive)",
    "alpha[2]"  = "Poverty (extensive)",
    "beta[2]"   = "Poverty (intensive)",
    "log_kappa"  = "log(kappa)"
  )

  trace_data <- do.call(rbind, lapply(key_params, function(p) {
    data.frame(
      iteration = draws$.iteration,
      chain     = factor(draws$.chain),
      value     = draws[[p]],
      parameter = param_display[p],
      stringsAsFactors = FALSE
    )
  }))
  trace_data$parameter <- factor(trace_data$parameter,
                                 levels = unname(param_display))

  ggplot(trace_data, aes(x = iteration, y = value, colour = chain)) +
    geom_line(alpha = 0.5, linewidth = 0.3) +
    facet_wrap(~ parameter, scales = "free_y", ncol = 2) +
    scale_colour_manual(
      values = c("1" = pal["extensive"], "2" = pal["intensive"],
                 "3" = pal["reference"], "4" = pal["dispersion"]),
      name = "Chain"
    ) +
    labs(x = "Iteration (post-warmup)", y = "Parameter value",
         title = "MCMC Trace Plots") +
    theme_minimal(base_size = 12) +
    theme(legend.position = "bottom",
          strip.text = element_text(face = "bold", size = 10))
}
```

For a more formal check, the summary output reports $\hat{R}$ (the potential
scale reduction factor) and effective sample size (ESS) for every parameter.
Values of $\hat{R}$ close to 1.00 and ESS well above 400 indicate reliable
inference.


## Posterior Predictive Checks

Convergence tells us the sampler is working, but not whether the model *fits
the data*. Posterior predictive checks (PPC) compare replicated data from the
posterior to the observations on two summary statistics:

1. **Zero rate** --- the fraction of providers with zero IT enrolment. Targets
   the extensive margin.
2. **Mean IT share** --- the average `y / n` among servers. Targets the
   intensive margin.

```{r load-ppc, eval = has_data}
ppc_obj <- readRDS(find_extdata("vig01_ppc.rds"))
ppc_obj
```

```{r fig-ppc, eval = has_data, fig.height = 4, fig.cap = "Posterior predictive checks. Left: replicated zero rates with the observed value (dashed line). Right: replicated mean IT shares. Both observed values fall within the predictive distributions."}
if (requireNamespace("ggplot2", quietly = TRUE)) {
  library(ggplot2)

  df_zr <- data.frame(value = ppc_obj$predicted$zero_rate$draws)
  df_it <- data.frame(value = ppc_obj$predicted$it_share$draws)
  obs_zr <- ppc_obj$observed$zero_rate
  obs_it <- ppc_obj$observed$it_share
  zr_pass <- ppc_obj$coverage$zero_rate$in_ci
  it_pass <- ppc_obj$coverage$it_share$in_ci

  p_zr <- ggplot(df_zr, aes(x = value)) +
    geom_histogram(bins = 40, fill = pal["extensive"], colour = "white",
                   linewidth = 0.2, alpha = 0.75) +
    geom_vline(xintercept = obs_zr, linetype = "dashed",
               colour = pal["extensive"], linewidth = 1) +
    annotate("text", x = obs_zr, y = Inf, vjust = -0.5,
             label = ifelse(zr_pass, "PASS", "FAIL"),
             colour = ifelse(zr_pass, pal["reference"], "red"),
             fontface = "bold", size = 5) +
    labs(x = "Replicated zero rate", y = "Count",
         subtitle = "Zero Rate (extensive)") +
    theme_minimal(base_size = 12) +
    theme(panel.grid.minor = element_blank())

  p_it <- ggplot(df_it, aes(x = value)) +
    geom_histogram(bins = 40, fill = pal["intensive"], colour = "white",
                   linewidth = 0.2, alpha = 0.75) +
    geom_vline(xintercept = obs_it, linetype = "dashed",
               colour = pal["intensive"], linewidth = 1) +
    annotate("text", x = obs_it, y = Inf, vjust = -0.5,
             label = ifelse(it_pass, "PASS", "FAIL"),
             colour = ifelse(it_pass, pal["reference"], "red"),
             fontface = "bold", size = 5) +
    labs(x = "Replicated mean IT share", y = "Count",
         subtitle = "IT Share (intensive)") +
    theme_minimal(base_size = 12) +
    theme(panel.grid.minor = element_blank())

  if (requireNamespace("patchwork", quietly = TRUE)) {
    library(patchwork)
    p_zr + p_it +
      plot_annotation(title = "Posterior Predictive Checks",
                      theme = theme(plot.title = element_text(size = 14,
                                                              face = "bold")))
  } else {
    print(p_zr)
    print(p_it)
  }
}
```

Both observed statistics fall comfortably within their respective 95%
predictive intervals, confirming that the model captures the zero-inflation
and the intensity distribution simultaneously.


## Cross-Validation

To evaluate *out-of-sample* predictive accuracy, we use Pareto-smoothed
importance sampling LOO (PSIS-LOO):

```{r load-loo, eval = has_data}
loo_obj <- readRDS(find_extdata("vig01_loo.rds"))
print(loo_obj)
```

Two quantities matter:

- **ELPD** (expected log predictive density): higher (less negative) is
  better. Most useful when *comparing* two models.
- **Pareto $k$ diagnostics**: observations with $k > 0.7$ deserve scrutiny;
  many above this threshold indicate unreliable LOO approximation.


## Fitted Values and Residuals

The `fitted()` function offers three types:

| Type          | Returns                                                    |
|:--------------|:-----------------------------------------------------------|
| `"response"`  | Overall expected IT count: $n_i \cdot q_i \cdot \mu_i$    |
| `"extensive"` | Participation probability $\hat{q}_i$                      |
| `"intensive"` | Conditional IT share $\hat{\mu}_i$                         |

Similarly, `residuals()` supports `"response"` (raw) and `"pearson"`
(standardised by the model-implied variance):

```{r fitted-head, eval = has_data}
cat("Fitted response (first 6):\n")
round(head(results$fitted_response), 3)
cat("\nPearson residuals (first 6):\n")
round(head(results$residuals_pearson), 3)
```

```{r fig-residuals, eval = has_data, fig.height = 4.5, fig.cap = "Residual diagnostics. Left: Pearson residuals versus fitted values with a loess smooth. Right: QQ plot comparing Pearson residuals to normal quantiles."}
if (requireNamespace("ggplot2", quietly = TRUE)) {
  library(ggplot2)

  resid_df <- data.frame(
    fitted  = results$fitted_response,
    pearson = results$residuals_pearson
  )

  p_resid <- ggplot(resid_df, aes(x = fitted, y = pearson)) +
    geom_hline(yintercept = 0, linetype = "dashed",
               colour = pal["reference"], linewidth = 0.6) +
    geom_point(alpha = 0.3, size = 1.2, colour = "grey40") +
    geom_smooth(method = "loess", se = TRUE, colour = pal["reference"],
                fill = pal["reference"], alpha = 0.12, linewidth = 0.8) +
    labs(x = "Fitted values", y = "Pearson residuals",
         subtitle = "Residuals vs. fitted") +
    theme_minimal(base_size = 12) +
    theme(panel.grid.minor = element_blank())

  qq_data <- data.frame(
    theoretical = qnorm(ppoints(length(resid_df$pearson))),
    sample      = sort(resid_df$pearson)
  )

  p_qq <- ggplot(qq_data, aes(x = theoretical, y = sample)) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed",
                colour = pal["reference"], linewidth = 0.6) +
    geom_point(alpha = 0.3, size = 1.2, colour = "grey40") +
    labs(x = "Theoretical quantiles", y = "Sample quantiles",
         subtitle = "Normal QQ plot") +
    theme_minimal(base_size = 12) +
    theme(panel.grid.minor = element_blank())

  if (requireNamespace("patchwork", quietly = TRUE)) {
    library(patchwork)
    p_resid + p_qq +
      plot_annotation(title = "Residual Diagnostics",
                      theme = theme(plot.title = element_text(size = 14,
                                                              face = "bold")))
  } else {
    print(p_resid)
    print(p_qq)
  }
}
```

A flat loess smooth near zero indicates no systematic misfit. Some tail
departures in the QQ plot are typical for count data and do not indicate
a problem with the hurdle specification.


## What's Next?

This vignette covered the core workflow. The remaining vignettes go deeper:

| Vignette | What you will learn |
|:---------|:--------------------|
| *Survey Design* | Incorporating complex survey weights and sandwich standard errors |
| *State-Varying Coefficients* | Hierarchical model with state-level random slopes |
| *Policy Moderators* | Cross-level interactions explaining state variation |
| *Marginal Effects* | AME decomposition and the poverty reversal probability |
