---
title: "Survey Design Corrections"
author: "JoonHo Lee"
date: "`r Sys.Date()`"
description: >
  Incorporating complex survey weights into Hurdle Beta-Binomial models:
  weighted pseudo-posteriors, sandwich standard errors, design effect ratios,
  and Wald confidence intervals.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Survey Design Corrections}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
library(hurdlebb)
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width  = 7,
  fig.height = 5,
  fig.align  = "center",
  out.width  = "85%"
)
set.seed(42)

# Project colour palette
pal <- c(extensive = "#4393C3", intensive = "#D6604D",
         reference = "#1B7837", dispersion = "#762A83")

# Helper: find extdata files in installed package OR source tree
find_extdata <- function(filename) {
  path <- system.file("extdata", filename, package = "hurdlebb")
  if (nzchar(path) && file.exists(path)) return(path)
  path2 <- file.path("..", "inst", "extdata", filename)
  if (file.exists(path2)) return(path2)
  ""
}

# Guard: pre-computed results must exist for eval'd chunks
has_data <- nzchar(find_extdata("vig02_results.rds"))
```

## Overview

Most large-scale social surveys do not draw observations at random from the
population. Instead, they use **complex survey designs** --- stratification,
clustering, and unequal probability sampling --- that make data collection
feasible but introduce dependencies the analyst must account for. Ignoring the
survey design leads to two problems:

1. **Biased point estimates.** If some groups are oversampled, the raw sample
   means do not estimate population quantities without reweighting.
2. **Underestimated standard errors.** Clustering means that observations
   within the same primary sampling unit (PSU) are more similar than a simple
   random sample would produce. Naive standard errors can be far too small.

This vignette shows how **hurdlebb** addresses both problems. By the end you
will understand:

- How to pass survey design variables (`weights`, `stratum`, `psu`) to
  `hbb()`.
- Why the resulting **pseudo-posterior** targets population parameters but
  produces standard deviations that are unreliable for inference.
- How the **sandwich variance estimator** recovers design-consistent standard
  errors.
- How to read **design effect ratios** (DERs) and what they tell you.
- Why **Wald confidence intervals** should replace posterior intervals when
  analysing survey data.

We assume you have read the *Getting Started* vignette and are comfortable
with the basic `hbb()` workflow.


## The Survey Design

The synthetic dataset `nsece_synth_small` mimics the 2019 National Survey of
Early Care and Education (NSECE), a nationally representative study that uses
a **stratified, clustered** design:

- **Strata** group providers by geography and provider type before sampling.
  Stratification improves precision by ensuring every subpopulation is
  represented.
- **Primary sampling units (PSUs)** are geographic clusters (roughly counties)
  within each stratum. Providers inside the same PSU share local economic and
  regulatory conditions, creating positive intra-cluster correlation.
- **Sampling weights** equal the inverse of each provider's selection
  probability, adjusted for non-response. A provider with weight 50
  represents roughly 50 providers in the population.

```{r design-vars}
data("nsece_synth_small", package = "hurdlebb")
cat("Sample size:", nrow(nsece_synth_small), "\n")
cat("Strata:     ", length(unique(nsece_synth_small$stratum)), "\n")
cat("PSUs:       ", length(unique(nsece_synth_small$psu)), "\n")
cat("Weight range:", round(range(nsece_synth_small$weight), 1), "\n")
cat("Weight CV:  ", round(sd(nsece_synth_small$weight) /
                          mean(nsece_synth_small$weight), 2), "\n")
```

```{r fig-weight-dist, fig.height = 3.5, fig.cap = "Distribution of survey weights. The right skew indicates that some providers represent many more population units than others."}
if (requireNamespace("ggplot2", quietly = TRUE)) {
  library(ggplot2)
  w <- nsece_synth_small$weight

  ggplot(data.frame(weight = w), aes(x = weight)) +
    geom_histogram(bins = 40, fill = pal["reference"], colour = "white",
                   alpha = 0.85, linewidth = 0.3) +
    geom_vline(xintercept = mean(w), linetype = "dashed",
               colour = "grey30", linewidth = 0.7) +
    annotate("text", x = mean(w) * 1.5, y = Inf, vjust = 2,
             label = sprintf("mean = %.1f", mean(w)),
             size = 3.5, colour = "grey30") +
    labs(x = "Survey weight", y = "Count",
         title = "Survey Weight Distribution") +
    theme_minimal(base_size = 12) +
    theme(panel.grid.minor = element_blank())
}
```


## Fitting the Weighted Model

Incorporating the survey design into `hbb()` requires three additional
arguments --- `weights`, `stratum`, and `psu` --- each taking the name of the
corresponding column in your data frame:

```r
# This is the code you would run (takes ~2-5 minutes):
fit_w <- hbb(
  y | trials(n_trial) ~ poverty + urban,
  data    = nsece_synth_small,
  weights = "weight",
  stratum = "stratum",
  psu     = "psu",
  chains  = 4,
  seed    = 42
)
```

Behind the scenes, `hbb()` normalises the weights (so they sum to the sample
size $N$), then multiplies each provider's log-likelihood contribution by its
weight. The result is a **weighted pseudo-posterior** that shifts point
estimates toward the population-representative values.

```{r load-results, eval = has_data}
results  <- readRDS(find_extdata("vig02_results.rds"))
sandwich <- readRDS(find_extdata("vig02_sandwich.rds"))
wald_ci  <- readRDS(find_extdata("vig02_wald_ci.rds"))
loo_comp <- readRDS(find_extdata("vig02_loo_compare.rds"))
```

```{r print-output, eval = has_data}
cat(paste(results$print_text, collapse = "\n"))
```

Note the **Weighted** label --- this confirms that survey weights were active
during estimation.


## The Pseudo-Posterior

In a standard Bayesian analysis the posterior is
$$\pi(\theta \mid y) \propto \pi(\theta) \prod_{i=1}^{N} L_i(\theta).$$
The pseudo-log-likelihood replaces this with
$$\tilde{\ell}(\theta) = \sum_{i=1}^{N} \tilde{w}_i \, \log L_i(\theta),$$
where $\tilde{w}_i = w_i / \bar{w}$ are the normalised weights. Providers
representing more population units contribute proportionally more to the
likelihood surface.

This is exactly what we want for point estimation: the posterior means target
the finite-population parameters. However, the posterior **standard
deviations** are no longer trustworthy for inference. The weights distort the
curvature of the likelihood --- a provider with weight 50 acts as though it
were replicated 50 times, artificially inflating the effective sample size.
The pseudo-posterior becomes too concentrated.

**Key insight:** the pseudo-posterior gives good point estimates but
misleading uncertainty. We need the sandwich variance estimator to recover
reliable standard errors.


## Sandwich Standard Errors

The sandwich variance estimator has a long history in frequentist survey
statistics. **hurdlebb** adapts it to the pseudo-posterior setting following
Williams & Savitsky (2021):

$$V_{\text{sand}} = H^{-1} \, J_{\text{cluster}} \, H^{-1}$$

where:

- $H$ is the **observed Fisher information** (the negative Hessian of the
  pseudo-log-posterior). Its inverse $H^{-1}$ is the "bread" --- the variance
  you would get with a simple random sample.
- $J_{\text{cluster}}$ is the **clustered outer-product-of-gradients** ---
  the "meat" of the sandwich. It sums score contributions at the PSU level,
  capturing intra-cluster correlation.

One function call computes everything:

```r
# In a live session:
sandwich <- sandwich_variance(fit_w)
```

The returned `hbb_sandwich` object contains the full matrix decomposition:

```{r sandwich-components, eval = has_data}
diag_df <- data.frame(
  parameter = sandwich$param_labels,
  H_inv     = round(diag(sandwich$H_obs_inv), 5),
  V_sand    = round(diag(sandwich$V_sand), 5)
)
diag_df$ratio <- round(diag_df$V_sand / diag_df$H_inv, 2)
diag_df
```

The `ratio` column is the design effect ratio (DER), which we examine next.


## Design Effect Ratios

The **design effect ratio** (DER) quantifies how much the survey design
inflates each parameter's variance relative to a simple random sample:

$$\text{DER}_p = \frac{V_{\text{sand},pp}}{H^{-1}_{pp}}.$$

A DER of 1.0 means the survey design has no impact. A DER of 5.0 means the
true variance is five times larger than a naive analysis would suggest ---
equivalently, the naive standard error is $\sqrt{5} \approx 2.24$ times too
small.

```{r der-table, eval = has_data}
der_df <- data.frame(
  parameter = sandwich$param_labels,
  DER       = round(sandwich$DER, 2),
  SE_ratio  = round(sqrt(sandwich$DER), 2)
)
knitr::kable(der_df, align = c("l", "r", "r"),
             caption = "Design effect ratios by parameter.")
```

```{r fig-der, eval = has_data, fig.height = 4, fig.cap = "Design effect ratios (DER) by parameter. The dashed line at DER = 1 marks no design effect. Intercepts show the greatest inflation; the dispersion parameter is largely unaffected."}
if (requireNamespace("ggplot2", quietly = TRUE)) {
  library(ggplot2)
  der_plot <- data.frame(
    parameter = factor(sandwich$param_labels,
                       levels = sandwich$param_labels[order(sandwich$DER)]),
    DER       = sandwich$DER,
    margin    = ifelse(grepl("^alpha", sandwich$param_labels), "Extensive",
                ifelse(grepl("^beta",  sandwich$param_labels), "Intensive",
                       "Dispersion"))
  )
  der_plot$margin <- factor(der_plot$margin,
                            levels = c("Extensive", "Intensive", "Dispersion"))

  ggplot(der_plot, aes(x = DER, y = parameter, fill = margin)) +
    geom_col(alpha = 0.85, width = 0.65) +
    geom_vline(xintercept = 1, linetype = "dashed", colour = "grey40",
               linewidth = 0.6) +
    geom_text(aes(label = sprintf("%.2f", DER)),
              hjust = -0.15, size = 3.3, colour = "grey20") +
    scale_fill_manual(
      values = c(Extensive = pal["extensive"], Intensive = pal["intensive"],
                 Dispersion = pal["dispersion"]),
      name = "Margin"
    ) +
    scale_x_continuous(expand = expansion(mult = c(0, 0.15))) +
    labs(x = "Design Effect Ratio (DER)", y = NULL,
         title = "Design Effect Ratios") +
    theme_minimal(base_size = 12) +
    theme(panel.grid.major.y = element_blank(),
          panel.grid.minor = element_blank(),
          legend.position = "bottom")
}
```

Several patterns are worth noting:

- **Intercepts have the largest DERs** (6.84 for extensive, 6.22 for
  intensive). This is common: the overall mean is most sensitive to clustering
  because every unit within a cluster shares the same local baseline.
- **Slope coefficients have moderate DERs** (1.38 to 4.65). Within-cluster
  variation in covariates partially offsets the clustering effect.
- **The dispersion parameter** ($\log\kappa$) has the smallest DER (1.17),
  because overdispersion is driven by within-unit variability, not clustering.

The mean DER across all seven parameters is 3.62. As a guideline: DERs above
2--3 indicate that ignoring the survey design would lead to materially
overconfident inference.


## Posterior vs. Wald Confidence Intervals

We can now contrast two types of intervals:

1. **Posterior intervals** (from the pseudo-posterior MCMC draws). These are
   too narrow because they do not account for the survey design.
2. **Wald intervals** constructed as $\hat\theta_p \pm z_{0.975} \times
   \text{SE}_{\text{sand},p}$. These are design-consistent.

The `summary()` method accepts a `sandwich` argument to produce the Wald
version. Compare the two:

```{r summary-posterior, eval = has_data}
cat("=== Posterior inference ===\n")
cat(paste(results$summary_text, collapse = "\n"))
```

```{r summary-wald, eval = has_data}
cat("=== Wald inference (sandwich SE) ===\n")
cat(paste(results$summary_wald_text, collapse = "\n"))
```

The Wald CI table from `compute_wald_ci()`:

```{r wald-table, eval = has_data}
knitr::kable(
  wald_ci[, c("parameter", "post_mean", "se", "z_stat",
              "p_value", "ci_lo", "ci_hi", "significant")],
  digits = 3, align = c("l", rep("r", 7)),
  caption = "Wald 95% confidence intervals from sandwich variance."
)
```

The forest plot below makes the comparison visual. For each parameter, the
thin line shows the posterior credible interval and the thick translucent line
shows the wider Wald confidence interval:

```{r fig-forest, eval = has_data, fig.height = 5, fig.cap = "Forest plot comparing posterior (thin lines, open circles) and Wald (thick translucent lines, filled triangles) 95% intervals. Wald intervals are consistently wider, reflecting the true sampling variability from the complex survey design."}
if (requireNamespace("ggplot2", quietly = TRUE)) {
  library(ggplot2)

  post_fe <- results$summary_obj$fixed_effects
  wald_fe <- results$summary_wald_obj$fixed_effects

  ci_df <- data.frame(
    parameter  = post_fe$parameter,
    estimate   = post_fe$estimate,
    post_lo    = post_fe$ci_lower,
    post_hi    = post_fe$ci_upper,
    wald_lo    = wald_fe$ci_lower,
    wald_hi    = wald_fe$ci_upper,
    margin     = ifelse(grepl("^alpha", post_fe$parameter), "Extensive",
                 ifelse(grepl("^beta",  post_fe$parameter), "Intensive",
                        "Dispersion")),
    stringsAsFactors = FALSE
  )
  ci_df$margin <- factor(ci_df$margin,
                         levels = c("Extensive", "Intensive", "Dispersion"))
  ci_df$ypos <- rev(seq_len(nrow(ci_df)))

  margin_col <- c(Extensive = pal["extensive"], Intensive = pal["intensive"],
                  Dispersion = pal["dispersion"])

  ggplot(ci_df) +
    # Wald CI (thick, behind)
    geom_segment(aes(x = wald_lo, xend = wald_hi,
                     y = ypos - 0.12, yend = ypos - 0.12,
                     colour = margin),
                 linewidth = 3, alpha = 0.3) +
    geom_point(aes(x = estimate, y = ypos - 0.12, colour = margin),
               shape = 17, size = 2.5) +
    # Posterior CI (thin, in front)
    geom_segment(aes(x = post_lo, xend = post_hi,
                     y = ypos + 0.12, yend = ypos + 0.12,
                     colour = margin),
                 linewidth = 0.9) +
    geom_point(aes(x = estimate, y = ypos + 0.12, colour = margin),
               shape = 1, size = 2.5, stroke = 0.8) +
    # Zero reference
    geom_vline(xintercept = 0, linetype = "dashed", colour = "grey50",
               linewidth = 0.5) +
    scale_colour_manual(values = margin_col, name = "Margin") +
    scale_y_continuous(breaks = ci_df$ypos, labels = ci_df$parameter) +
    labs(x = "Estimate (95% interval)", y = NULL,
         title = "Posterior vs. Wald Confidence Intervals",
         subtitle = "Open circles / thin = Posterior;  Filled triangles / thick = Wald") +
    theme_minimal(base_size = 12) +
    theme(legend.position = "bottom",
          panel.grid.major.y = element_blank(),
          panel.grid.minor = element_blank(),
          plot.subtitle = element_text(size = 9, colour = "grey40"))
}
```

The pattern is striking. For `alpha_poverty`, the posterior standard error is
0.089 while the sandwich standard error is 0.191 --- more than twice as large,
consistent with a DER of 4.65 ($\sqrt{4.65} \approx 2.16$). Using the
posterior interval alone, you might conclude that poverty has a statistically
significant negative effect on participation. The Wald interval, which properly
accounts for the survey design, is much wider and includes zero.


## CI Width Comparison

A grouped bar chart makes the inflation tangible:

```{r fig-ci-width, eval = has_data, fig.height = 4.5, fig.cap = "CI width comparison. Wald intervals (red) are uniformly wider than posterior intervals (blue). The ratio printed above each pair approximates the square root of the DER."}
if (requireNamespace("ggplot2", quietly = TRUE)) {
  library(ggplot2)

  post_fe <- results$summary_obj$fixed_effects
  wald_fe <- results$summary_wald_obj$fixed_effects

  width_df <- data.frame(
    parameter  = post_fe$parameter,
    post_width = post_fe$ci_upper - post_fe$ci_lower,
    wald_width = wald_fe$ci_upper - wald_fe$ci_lower,
    stringsAsFactors = FALSE
  )
  width_df$ratio <- round(width_df$wald_width / width_df$post_width, 2)

  width_long <- rbind(
    data.frame(parameter = width_df$parameter,
               width = width_df$post_width, type = "Posterior"),
    data.frame(parameter = width_df$parameter,
               width = width_df$wald_width, type = "Wald (sandwich)")
  )
  width_long$parameter <- factor(width_long$parameter,
                                 levels = width_df$parameter)
  width_long$type <- factor(width_long$type,
                            levels = c("Posterior", "Wald (sandwich)"))

  ggplot(width_long, aes(x = parameter, y = width, fill = type)) +
    geom_col(position = position_dodge(width = 0.7), width = 0.6) +
    geom_text(data = width_df,
              aes(x = parameter, y = wald_width,
                  label = paste0(ratio, "x")),
              inherit.aes = FALSE,
              vjust = -0.5, size = 3, colour = "grey30") +
    scale_fill_manual(
      values = c("Posterior" = pal["extensive"],
                 "Wald (sandwich)" = pal["intensive"]),
      name = NULL
    ) +
    labs(x = NULL, y = "95% CI Width",
         title = "Posterior vs. Wald Interval Widths") +
    theme_minimal(base_size = 12) +
    theme(axis.text.x = element_text(angle = 30, hjust = 1),
          legend.position = "top",
          panel.grid.minor = element_blank())
}
```


## Model Comparison with LOO

You may wonder whether the weighted model fits better than the unweighted
version. LOO-CV provides the comparison:

```r
# In a live session:
loo_comp <- hbb_loo_compare(base = fit_base, weighted = fit_w)
```

```{r loo-comparison, eval = has_data}
cat("Model comparison:\n")
print(loo_comp$comparison[, c("model", "elpd_loo", "elpd_diff", "se_diff")])
cat("\nPairwise z-ratio:\n")
print(loo_comp$pairwise)
```

The base (unweighted) model has a slightly higher ELPD ($-1485.8$) than the
weighted model ($-1496.9$), with a difference of $11.07$ and $z = 1.91$.
This difference is **not significant** at conventional levels.

This result may seem counterintuitive --- if weights improve inference, why
does the weighted model predict less well? The answer lies in what LOO
measures. LOO evaluates **predictive accuracy for the sample at hand**, not
population inference. The survey weights distort the likelihood to target
population parameters, which can reduce in-sample predictive performance.
This is expected behaviour.

**Guideline:** Do not use LOO to choose between weighted and unweighted models.
The decision should be driven by your inferential goal: if the data come from a
complex survey and you want population-level conclusions, use weights.


## Complete Workflow Summary

Putting it all together:

```r
# 1. Fit weighted model
fit <- hbb(
  y | trials(n_trial) ~ poverty + urban,
  data = nsece_synth_small,
  weights = "weight", stratum = "stratum", psu = "psu",
  chains = 4, seed = 42
)

# 2. Posterior summary (naive SEs --- useful for MCMC diagnostics)
summary(fit)

# 3. Sandwich variance (design-corrected SEs)
sand <- sandwich_variance(fit)

# 4. Wald summary (design-corrected CIs + DER)
summary(fit, sandwich = sand)

# 5. Explicit Wald CI table
wald <- compute_wald_ci(coef(fit), sand$V_sand, level = 0.95)
```


## When to Use Corrections

| Situation | Recommendation |
|:----------|:---------------|
| Simple random sample | Use unweighted `hbb()`, posterior CIs |
| Stratified design, no clustering | Weights may suffice; check DER |
| Stratified clustered design | Full pipeline: weights + sandwich + Wald CIs |
| DER $< 1.2$ for all parameters | Posterior CIs are adequate |
| DER $> 2$ for key parameters | Sandwich-based Wald CIs are essential |

**Reporting.** In your manuscript, report posterior means as point estimates
and Wald confidence intervals as the uncertainty measure. Include the DER
table in a supplement so readers can gauge the severity of design effects.


## What's Next?

| Vignette | What you will learn |
|:---------|:--------------------|
| *State-Varying Coefficients* | Hierarchical model with state-level random slopes + survey weights |
| *Policy Moderators* | Cross-level interactions explaining state variation |
| *Marginal Effects* | AME decomposition with design-corrected standard errors |
