---
title: "State-Varying Coefficients"
author: "JoonHo Lee"
date: "`r Sys.Date()`"
description: >
  Extending the Hurdle Beta-Binomial to capture state-level heterogeneity
  through random intercepts and slopes, with interpretation of variance
  components, cross-margin correlations, and model comparison via LOO-CV.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{State-Varying Coefficients}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
library(hurdlebb)
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width  = 7,
  fig.height = 5,
  fig.align  = "center",
  out.width  = "85%"
)
set.seed(42)

# Project colour palette
pal <- c(extensive = "#4393C3", intensive = "#D6604D",
         reference = "#1B7837", dispersion = "#762A83")

# Helper: find extdata files in installed package OR source tree
find_extdata <- function(filename) {
  path <- system.file("extdata", filename, package = "hurdlebb")
  if (nzchar(path) && file.exists(path)) return(path)
  path2 <- file.path("..", "inst", "extdata", filename)
  if (file.exists(path2)) return(path2)
  ""
}

# Guard: pre-computed results must exist for eval'd chunks
has_data <- nzchar(find_extdata("vig03_results.rds"))
```

## Overview

The previous vignettes treated the relationship between covariates and IT
enrolment as constant across all 51 states. That assumption is convenient but
unrealistic. States differ in childcare regulations, subsidy structures,
labour markets, and demographics. If these differences are systematic,
pooling all states together can mask important heterogeneity --- or produce a
"national average" that describes no state accurately.

**State-varying coefficients (SVC)** address this by allowing each state to
have its own regression coefficients for both margins. Instead of a single
poverty slope, we estimate a global average plus a state-specific deviation.
States with more data pull toward local evidence; states with sparse data are
partially pooled toward the national mean.

In this vignette you will learn how to:

1. **Understand** the hierarchical model structure behind SVC.
2. **Specify** an SVC model using random-effects formula syntax in `hbb()`.
3. **Interpret** fixed effects alongside variance components.
4. **Examine** random-effect standard deviations ($\tau$).
5. **Read** the cross-margin correlation matrix.
6. **Visualise** state-level deviations with caterpillar plots.
7. **Compare** the SVC model to the base model via LOO-CV.

We assume familiarity with the basic `hbb()` workflow from the *Getting
Started* vignette.


## The Hierarchical Structure

The state-varying coefficient model extends the base Hurdle Beta-Binomial by
adding state-level random effects to both margins. Let $s(i)$ denote the
state of provider $i$. The linear predictors become:

**Extensive margin** (participation):
$$\text{logit}(q_i) = X_i'\alpha + X_i'\delta_s^{(\alpha)}$$

**Intensive margin** (enrolment share):
$$\text{logit}(\mu_i) = X_i'\beta + X_i'\delta_s^{(\beta)}$$

Here $\alpha$ and $\beta$ are **global fixed effects** --- the
national-average coefficients. The vectors
$\delta_s = (\delta_s^{(\alpha)}, \delta_s^{(\beta)})$ are **state-level
deviations** that shift each state's coefficients away from the national
mean. The effective coefficient for state $s$ is:

$$\alpha_s = \alpha + \delta_s^{(\alpha)}, \qquad
  \beta_s = \beta + \delta_s^{(\beta)}.$$

The random effects follow a multivariate normal distribution:

$$\delta_s \sim \text{MVN}(0, \Sigma), \qquad s = 1, \ldots, S,$$

where $\Sigma$ is a $2P \times 2P$ covariance matrix ($P$ = number of
covariates including the intercept). The **diagonal** of $\Sigma$ tells us
how much each coefficient varies across states. The **off-diagonal** elements
capture correlations, including **cross-margin correlations** --- for
instance, whether states where poverty encourages participation are the same
states where poverty raises intensity.

To ensure positive-definiteness, the model uses a Cholesky decomposition:
$\Sigma = \text{diag}(\tau) \, L L' \, \text{diag}(\tau)$, where $\tau$
contains the standard deviations and $L$ is a lower-triangular Cholesky
factor with an LKJ prior. This is the standard approach in Stan and brms.


## The Data

We use the full synthetic NSECE dataset, which has enough observations per
state to support random slopes.

```{r load-data}
data("nsece_synth", package = "hurdlebb")
cat("Sample size:", nrow(nsece_synth), "\n")
cat("States:     ", length(unique(nsece_synth$state_id)), "\n")
cat("Obs/state:  ", round(nrow(nsece_synth) /
                          length(unique(nsece_synth$state_id)), 1), "\n")
cat("Zero rate:  ", round(1 - mean(nsece_synth$z), 3), "\n")
```

With approximately 133 observations per state on average, there is adequate
information to estimate state-specific intercepts and slopes. The smaller
`nsece_synth_small` (approximately 500 rows, roughly 10 per state) is not
suitable for SVC models because the random effects would be dominated by
the prior.


## Specifying the SVC Model

In **hurdlebb**, random effects are added to the formula using the familiar
`(slopes | group)` syntax. To allow both the intercept and the slope
coefficients for poverty and urban to vary by state:

```r
# This is the code you would run (takes ~45-60 minutes on full data):
fit_svc <- hbb(
  y | trials(n_trial) ~ poverty + urban + (poverty + urban | state_id),
  data    = nsece_synth,
  weights = "weight",
  stratum = "stratum",
  psu     = "psu",
  chains  = 4,
  seed    = 42
)
```

Several things to note:

- **Random effects.** The term `(poverty + urban | state_id)` adds random
  intercepts and random slopes for both covariates, on both margins
  simultaneously. This produces $2P = 6$ random effects per state (three for
  the extensive margin, three for the intensive margin).
- **Survey weights.** We include `weights`, `stratum`, and `psu` as in the
  survey design vignette. The SVC and survey-weight features compose cleanly:
  the model uses the `hbb_svc_weighted` Stan variant.
- **Runtime.** The SVC model is substantially more expensive than the base
  model because it must estimate $51 \times 6 = 306$ random effects plus
  their covariance structure. Expect 45--60 minutes on a modern workstation.


## Loading Pre-computed Results

As with previous vignettes, we load saved output:

```{r load-results, eval = has_data}
results <- readRDS(find_extdata("vig03_results.rds"))
```

The `print()` method gives a compact overview:

```{r show-print, eval = has_data}
cat(paste(results$print_text, collapse = "\n"))
```


## Model Summary

The `summary()` method reports posterior means, standard deviations, and 95%
credible intervals for the fixed effects, plus the variance components:

```{r show-summary, eval = has_data}
cat(paste(results$summary_text, collapse = "\n"))
```

### Reading the fixed-effects table

The fixed effects represent **national-average** coefficients --- the typical
relationship between each covariate and IT enrolment across all states.

```{r coef-table, eval = has_data}
results$coef_both
```

The poverty reversal persists at the national level:

```{r poverty-reversal, eval = has_data}
cat(sprintf("Extensive margin (alpha_poverty): %+.3f\n",
            results$coef_both["alpha_poverty"]))
cat(sprintf("Intensive margin  (beta_poverty): %+.3f\n",
            results$coef_both["beta_poverty"]))
```

Higher neighbourhood poverty is associated with lower participation
probability (negative $\alpha$) but higher enrolment share among servers
(positive $\beta$). The key question the SVC model answers is: *does this
reversal hold uniformly across states, or do some states deviate from the
pattern?*


## Global Fixed Effects Forest Plot

With random effects in the model, the fixed effects $(\alpha, \beta)$
represent **population-average** coefficients. Their credible intervals are
wider than in the pooled model because the random effects absorb much of the
covariate-outcome variation that previously sharpened the fixed-effect
posteriors. The forest plot below makes this visible:

```{r fig-forest-global, eval = has_data, fig.height = 5, fig.cap = "Global fixed effects (population-average) with 95% credible intervals. Blue = extensive margin, red = intensive margin, purple = dispersion. Wider intervals compared to the pooled model reflect state heterogeneity captured by random effects. The poverty signs persist, confirming the reversal is a robust national phenomenon."}
if (requireNamespace("ggplot2", quietly = TRUE)) {
  library(ggplot2)

  fe <- results$summary_obj$fixed_effects

  # Assign margin groups
  fe$margin_group <- ifelse(grepl("^alpha", fe$parameter), "Extensive",
                    ifelse(grepl("^beta",  fe$parameter), "Intensive",
                           "Dispersion"))
  fe$margin_group <- factor(fe$margin_group,
                            levels = c("Extensive", "Intensive", "Dispersion"))

  # Clean labels with Greek letters
  fe$label <- gsub("alpha_", "\u03b1 : ", fe$parameter)
  fe$label <- gsub("beta_",  "\u03b2 : ", fe$label)
  fe$label <- gsub("\\(Intercept\\)", "Intercept", fe$label)
  fe$label <- gsub("log_kappa", "log(\u03ba)", fe$label)
  fe$label <- factor(fe$label, levels = rev(fe$label))

  ggplot(fe, aes(x = estimate, y = label, colour = margin_group)) +
    geom_vline(xintercept = 0, linetype = "dashed",
               colour = pal["reference"], linewidth = 0.5) +
    geom_pointrange(aes(xmin = ci_lower, xmax = ci_upper),
                    size = 0.5, linewidth = 0.7, fatten = 2.5) +
    # Annotate point estimates
    geom_text(aes(label = sprintf("%.2f", estimate)),
              hjust = -0.3, vjust = -0.8, size = 2.8,
              show.legend = FALSE) +
    scale_colour_manual(
      values = c(Extensive  = pal["extensive"],
                 Intensive  = pal["intensive"],
                 Dispersion = pal["dispersion"]),
      name = "Margin"
    ) +
    labs(x = "Posterior estimate", y = NULL,
         title = "Global Fixed Effects (Population-Average Coefficients)",
         subtitle = paste("SVC model: state-level heterogeneity",
                          "absorbed into random effects")) +
    theme_minimal(base_size = 12) +
    theme(legend.position = "bottom",
          panel.grid.minor = element_blank(),
          panel.grid.major.y = element_blank(),
          plot.subtitle = element_text(size = 9, colour = "grey45"))
}
```

Compared to the pooled model in the *Getting Started* vignette, the slope
intervals are substantially wider because state-level variation is now
modelled explicitly. The poverty signs persist ($\alpha < 0$, $\beta > 0$),
confirming the reversal is a robust population-level phenomenon, not an
artefact of ignoring state variation. The log-dispersion $\log\kappa$ remains
tightly estimated because it does not receive random effects.


## Random-Effect Standard Deviations ($\tau$)

The $\tau$ parameters measure how much each coefficient varies across states.
Larger $\tau$ means more between-state heterogeneity for that coefficient.

```{r tau-table, eval = has_data}
tau <- results$tau_table
knitr::kable(
  tau[, c("parameter", "margin", "covariate", "estimate", "sd",
          "ci_lower", "ci_upper", "rhat", "ess_bulk")],
  digits  = 3,
  align   = c("l", "l", "l", rep("r", 6)),
  caption = "Random-effect standard deviations by parameter."
)
```

```{r fig-tau, eval = has_data, fig.height = 4.5, fig.cap = "Random-effect standard deviations (tau) with 95% credible intervals. The extensive-margin intercept dominates cross-state variation. Slope coefficients vary modestly, suggesting the direction of covariate effects is relatively consistent nationwide. Blue = extensive margin, red = intensive margin."}
if (requireNamespace("ggplot2", quietly = TRUE)) {
  library(ggplot2)

  tau_df <- results$tau_table

  # Create informative labels with Greek letters
  tau_df$label <- paste0(
    ifelse(tau_df$margin == "extensive", "\u03b1", "\u03b2"),
    " : ", tau_df$covariate
  )

  # Capitalise margin for legend
  tau_df$Margin <- ifelse(tau_df$margin == "extensive",
                          "Extensive", "Intensive")

  # Sort by estimate so largest tau appears at top
  tau_df <- tau_df[order(tau_df$estimate), ]
  tau_df$label <- factor(tau_df$label, levels = tau_df$label)

  ggplot(tau_df, aes(x = estimate, y = label, colour = Margin)) +
    geom_vline(xintercept = 0, linetype = "dotted", colour = "grey60",
               linewidth = 0.4) +
    # CI bars as thick translucent background
    geom_linerange(aes(xmin = ci_lower, xmax = ci_upper),
                   linewidth = 2.5, alpha = 0.25) +
    # Point estimates
    geom_point(size = 3) +
    # Numeric labels
    geom_text(aes(label = sprintf("%.3f", estimate)),
              hjust = -0.4, size = 3.2, show.legend = FALSE) +
    scale_colour_manual(
      values = c(Extensive = pal["extensive"],
                 Intensive = pal["intensive"]),
      name = "Margin"
    ) +
    scale_x_continuous(expand = expansion(mult = c(0.02, 0.18))) +
    labs(x = expression(tau ~ "(random-effect SD)"),
         y = NULL,
         title = "Cross-State Variation in Coefficients") +
    theme_minimal(base_size = 12) +
    theme(legend.position = "bottom",
          panel.grid.minor = element_blank(),
          panel.grid.major.y = element_blank())
}
```

Several patterns are worth highlighting:

- **Intercept heterogeneity is largest.** The $\tau$ values for the intercept
  terms are the largest, meaning that states differ most in their baseline
  rates of IT participation and IT share. This is intuitive: different states
  have very different childcare landscapes that cannot be captured by the
  observed covariates alone.
- **Slope heterogeneity is smaller but non-trivial.** The poverty and urban
  slopes also vary across states, suggesting that the same change in
  neighbourhood poverty has different implications depending on the state.
  This is the substantive payoff of the SVC model.
- **Convergence diagnostics.** All $\hat{R}$ values are close to 1.00
  and ESS values are adequate for reliable inference.


## Cross-Margin Correlation Matrix

The correlation matrix of the random effects reveals how state-level
deviations co-vary across parameters and margins. This is one of the most
informative aspects of the SVC model.

```{r corr-matrix, eval = has_data}
R_mat <- results$corr_matrix
knitr::kable(
  round(R_mat, 2),
  caption = "Correlation matrix of state-level random effects."
)
```

```{r fig-corr, eval = has_data, fig.height = 5.5, fig.width = 6.5, fig.cap = "Heatmap of the 6x6 random-effect correlation matrix. Diverging colour scale: blue = negative, white = zero, red = positive. Heavy lines separate the extensive (left/top) and intensive (right/bottom) blocks. The off-diagonal blocks capture the cross-margin dependencies central to this model."}
if (requireNamespace("ggplot2", quietly = TRUE)) {
  library(ggplot2)

  R_mat <- results$corr_matrix
  param_names <- rownames(R_mat)
  n_par <- length(param_names)

  # Create short display labels with Greek letters
  short_labels <- gsub("alpha_", "\u03b1:", param_names)
  short_labels <- gsub("beta_",  "\u03b2:", short_labels)
  short_labels <- gsub("\\(Intercept\\)", "Int", short_labels)

  # Build long-format data
  corr_long <- expand.grid(
    row = seq_len(n_par),
    col = seq_len(n_par)
  )
  corr_long$value    <- as.vector(R_mat)
  corr_long$row_name <- factor(short_labels[corr_long$row],
                                levels = rev(short_labels))
  corr_long$col_name <- factor(short_labels[corr_long$col],
                                levels = short_labels)
  # Text colour: white on dark tiles, black on light tiles
  corr_long$text_col <- ifelse(abs(corr_long$value) > 0.5,
                                "white", "grey20")

  ggplot(corr_long, aes(x = col_name, y = row_name, fill = value)) +
    geom_tile(colour = "white", linewidth = 0.8) +
    geom_text(aes(label = sprintf("%.2f", value), colour = text_col),
              size = 3.2, show.legend = FALSE) +
    scale_fill_gradient2(
      low      = pal["extensive"],
      mid      = "white",
      high     = pal["intensive"],
      midpoint = 0,
      limits   = c(-1, 1),
      name     = "Correlation"
    ) +
    scale_colour_identity() +
    # Block-boundary lines separating extensive and intensive
    geom_hline(yintercept = 3.5, colour = "grey30", linewidth = 0.7) +
    geom_vline(xintercept = 3.5, colour = "grey30", linewidth = 0.7) +
    # Block labels below the axis
    annotate("text", x = 2, y = 0.3, label = "Extensive",
             size = 3.2, fontface = "bold", colour = pal["extensive"]) +
    annotate("text", x = 5, y = 0.3, label = "Intensive",
             size = 3.2, fontface = "bold", colour = pal["intensive"]) +
    coord_fixed(clip = "off") +
    labs(x = NULL, y = NULL,
         title = "Random-Effect Correlation Matrix") +
    theme_minimal(base_size = 12) +
    theme(panel.grid = element_blank(),
          axis.text.x = element_text(angle = 45, hjust = 1, size = 9),
          axis.text.y = element_text(size = 9),
          legend.position = "right",
          plot.margin = margin(5, 5, 25, 5))
}
```

### Interpreting cross-margin correlations

The most substantively interesting entries are the **cross-margin
correlations** --- the off-diagonal block linking $\alpha$ deviations to
$\beta$ deviations. A **positive** correlation between the intercepts means
that states with higher participation also show higher intensity. A
**negative** correlation suggests a tradeoff: broader participation comes
with lower average intensity. The cross-margin correlation for the poverty
slope reveals whether the poverty reversal is amplified or attenuated in
certain states.


## Caterpillar Plot: Poverty Effects Across States

The poverty coefficient is the substantive centrepiece. The caterpillar plot
below ranks states by their poverty random effect and shows 95% credible
intervals, with states whose intervals exclude zero highlighted in colour.

```{r fig-caterpillar-poverty, eval = has_data, fig.height = 6, fig.width = 8, fig.cap = "Caterpillar plots of state-level poverty random effects. Left: extensive margin (participation probability). Right: intensive margin (IT share among servers). States sorted by point estimate within each panel. Red points highlight states whose 95% CI excludes zero."}
if (requireNamespace("ggplot2", quietly = TRUE)) {
  library(ggplot2)

  re_df <- results$random_effects_df

  # Filter to poverty effects
  pov_re <- re_df[re_df$covariate == "poverty", ]
  pov_re$Margin <- ifelse(pov_re$margin == "extensive",
                          "Extensive (participation)",
                          "Intensive (IT share)")
  pov_re$Margin <- factor(pov_re$Margin,
                          levels = c("Extensive (participation)",
                                     "Intensive (IT share)"))

  # Sort states within each margin by estimate
  pov_re <- do.call(rbind, lapply(split(pov_re, pov_re$Margin), function(df) {
    df$rank <- rank(df$estimate, ties.method = "first")
    df
  }))

  # Flag states where CI excludes zero
  pov_re$significant <- (pov_re$ci_lower > 0) | (pov_re$ci_upper < 0)
  pov_re$sig_label   <- ifelse(pov_re$significant,
                               "CI excludes zero", "CI includes zero")

  ggplot(pov_re, aes(x = estimate, y = rank, colour = sig_label)) +
    geom_vline(xintercept = 0, linetype = "dashed",
               colour = pal["reference"], linewidth = 0.5) +
    geom_pointrange(aes(xmin = ci_lower, xmax = ci_upper),
                    size = 0.25, linewidth = 0.35, fatten = 1.8) +
    facet_wrap(~ Margin, scales = "free_x") +
    scale_colour_manual(
      values = c("CI excludes zero" = pal["intensive"],
                 "CI includes zero" = "grey55"),
      name = NULL
    ) +
    scale_y_continuous(breaks = seq(5, 50, by = 5),
                       minor_breaks = NULL) +
    labs(x = "State-level poverty random effect",
         y = "States (sorted by estimate)",
         title = "State-Varying Poverty Effects") +
    theme_minimal(base_size = 12) +
    theme(legend.position = "bottom",
          panel.grid.minor = element_blank(),
          panel.grid.major.y = element_line(colour = "grey92",
                                            linewidth = 0.3),
          strip.text = element_text(face = "bold", size = 11))
}
```

The caterpillar plots reveal that while most states cluster near zero (the
national average), a handful of states deviate noticeably. On the extensive
margin, some states show a distinctly more negative poverty effect (poverty is
a stronger barrier to IT service), while others are closer to neutral. The
intensive margin deviations tend to be smaller, consistent with the lower
$\tau_{\beta,\text{poverty}}$ observed earlier.


## Caterpillar Plot: All Covariates

For a comprehensive view, we examine all six random-effect dimensions ---
intercept, poverty, and urban on both margins --- in a single faceted display.
This reveals the overall heterogeneity landscape at a glance.

```{r fig-caterpillar-all, eval = has_data, fig.height = 8.5, fig.width = 9, fig.cap = "Caterpillar plots for all six state-varying coefficients. Rows: covariates (Intercept, Poverty, Urban). Columns: margins (Extensive, Intensive). States sorted by point estimate within each panel. The intercept panels show the widest spread, consistent with the tau estimates. Slope panels show tighter clustering near zero."}
if (requireNamespace("ggplot2", quietly = TRUE)) {
  library(ggplot2)

  re_df <- results$random_effects_df

  # Create readable facet labels
  re_df$Margin <- ifelse(re_df$margin == "extensive",
                         "Extensive", "Intensive")
  re_df$Margin <- factor(re_df$Margin, levels = c("Extensive", "Intensive"))
  re_df$Covariate <- factor(re_df$covariate,
                            levels = c("(Intercept)", "poverty", "urban"),
                            labels = c("Intercept", "Poverty", "Urban"))

  # Sort within each panel by estimate
  re_df <- do.call(rbind, lapply(
    split(re_df, interaction(re_df$Margin, re_df$Covariate, drop = TRUE)),
    function(df) {
      df$rank <- rank(df$estimate, ties.method = "first")
      df
    }
  ))

  # Flag significance
  re_df$significant <- (re_df$ci_lower > 0) | (re_df$ci_upper < 0)
  re_df$sig_label   <- ifelse(re_df$significant,
                              "CI excludes zero", "CI includes zero")

  ggplot(re_df, aes(x = estimate, y = rank, colour = sig_label)) +
    geom_vline(xintercept = 0, linetype = "dashed",
               colour = pal["reference"], linewidth = 0.4) +
    geom_pointrange(aes(xmin = ci_lower, xmax = ci_upper),
                    size = 0.18, linewidth = 0.28, fatten = 1.5) +
    facet_grid(Covariate ~ Margin, scales = "free") +
    scale_colour_manual(
      values = c("CI excludes zero" = pal["intensive"],
                 "CI includes zero" = "grey55"),
      name = NULL
    ) +
    scale_y_continuous(breaks = seq(10, 50, by = 10),
                       minor_breaks = NULL) +
    labs(x = "State-level random effect",
         y = "States (sorted by estimate)",
         title = "State-Varying Coefficients: All Parameters") +
    theme_minimal(base_size = 12) +
    theme(legend.position = "bottom",
          panel.grid.minor = element_blank(),
          panel.grid.major.y = element_line(colour = "grey92",
                                            linewidth = 0.3),
          strip.text = element_text(face = "bold", size = 10))
}
```

The intercept panels show the most dramatic variation, spanning roughly
$\pm 1$ on the extensive margin and $\pm 0.3$ on the intensive margin.
The slope panels display much tighter spreads. This hierarchy is exactly what
the $\tau$ estimates predict: the intercept SDs are an order of magnitude
larger than the slope SDs.


## Cross-Margin Scatter: Poverty Reversal at the State Level

A key question is whether the poverty reversal --- negative on the extensive
margin, positive on the intensive margin --- is a national-average phenomenon
or whether it manifests within individual states. We can examine this by
plotting each state's poverty random effect on the extensive margin against its
intensive-margin counterpart. Adding the global fixed effects determines
whether each state's *effective* coefficient exhibits the reversal pattern.

```{r fig-scatter-poverty, eval = has_data, fig.height = 5.5, fig.width = 6, fig.cap = "Cross-margin scatter of state-level poverty random effects. Each point is one state. Purple points indicate states whose effective coefficient (global + random) exhibits the full reversal pattern (negative extensive, positive intensive). Coloured dashed lines mark thresholds where the effective coefficient switches sign."}
if (requireNamespace("ggplot2", quietly = TRUE)) {
  library(ggplot2)

  re_df <- results$random_effects_df
  fe    <- results$summary_obj$fixed_effects

  # Extract poverty random effects for each margin
  pov_ext <- re_df[re_df$covariate == "poverty" & re_df$margin == "extensive",
                   c("state_id", "estimate")]
  names(pov_ext)[2] <- "delta_alpha"
  pov_int <- re_df[re_df$covariate == "poverty" & re_df$margin == "intensive",
                   c("state_id", "estimate")]
  names(pov_int)[2] <- "delta_beta"
  scatter_df <- merge(pov_ext, pov_int, by = "state_id")

  # Global fixed effects for poverty
  alpha_pov <- fe$estimate[fe$parameter == "alpha_poverty"]
  beta_pov  <- fe$estimate[fe$parameter == "beta_poverty"]

  # Effective state-level coefficients
  scatter_df$alpha_eff <- alpha_pov + scatter_df$delta_alpha
  scatter_df$beta_eff  <- beta_pov  + scatter_df$delta_beta

  # Reversal: alpha_eff < 0 AND beta_eff > 0
  scatter_df$pattern <- ifelse(
    scatter_df$alpha_eff < 0 & scatter_df$beta_eff > 0,
    "Reversal", "No reversal"
  )

  n_rev <- sum(scatter_df$pattern == "Reversal")
  n_tot <- nrow(scatter_df)

  # Axis ranges for annotation placement
  x_range <- range(scatter_df$delta_alpha)
  y_range <- range(scatter_df$delta_beta)

  ggplot(scatter_df, aes(x = delta_alpha, y = delta_beta,
                         colour = pattern)) +
    # Reference axes (zero in random-effect space)
    geom_hline(yintercept = 0, linetype = "dotted", colour = "grey75",
               linewidth = 0.4) +
    geom_vline(xintercept = 0, linetype = "dotted", colour = "grey75",
               linewidth = 0.4) +
    # Thresholds where effective coefficient = 0
    geom_vline(xintercept = -alpha_pov, linetype = "dashed",
               colour = pal["extensive"], linewidth = 0.6, alpha = 0.7) +
    geom_hline(yintercept = -beta_pov, linetype = "dashed",
               colour = pal["intensive"], linewidth = 0.6, alpha = 0.7) +
    # Points
    geom_point(size = 2.8, alpha = 0.85) +
    scale_colour_manual(
      values = c("Reversal"    = pal["dispersion"],
                 "No reversal" = "grey50"),
      name = NULL
    ) +
    # Count of reversal states
    annotate("text",
             x = x_range[1] + diff(x_range) * 0.02,
             y = y_range[2] - diff(y_range) * 0.02,
             label = sprintf("%d / %d states\nshow reversal", n_rev, n_tot),
             size = 3.8, fontface = "italic", colour = pal["dispersion"],
             lineheight = 0.9, hjust = 0, vjust = 1) +
    # Axis threshold labels
    annotate("text",
             x = -alpha_pov, y = y_range[1] - diff(y_range) * 0.08,
             label = expression(alpha[eff] == 0),
             size = 3, colour = pal["extensive"]) +
    annotate("text",
             x = x_range[2] + diff(x_range) * 0.05,
             y = -beta_pov,
             label = expression(beta[eff] == 0),
             size = 3, colour = pal["intensive"]) +
    labs(x = expression(delta[alpha] ~ "(extensive poverty random effect)"),
         y = expression(delta[beta]  ~ "(intensive poverty random effect)"),
         title = "Cross-Margin Poverty Effects by State") +
    theme_minimal(base_size = 12) +
    theme(legend.position = "bottom",
          panel.grid.minor = element_blank())
}
```

The dashed coloured lines mark the thresholds where the effective
state-level coefficient switches sign (i.e., $\alpha + \delta_\alpha = 0$
and $\beta + \delta_\beta = 0$). States in the upper-left quadrant relative
to these thresholds exhibit the full reversal pattern: poverty reduces the
probability of IT service but increases IT intensity among servers.

If most states fall in this quadrant, the poverty reversal is a genuine
national phenomenon rather than an artefact of a few outlier states. States
outside this quadrant merit substantive investigation --- they may have
policies that either mitigate the poverty barrier to participation or dampen
the poverty--intensity link.


## MCMC Diagnostics: Trace Plots for $\tau$ Parameters

The random-effect SDs ($\tau_1, \ldots, \tau_6$) are the hardest parameters
to estimate because they sit at the top of the hierarchy and are informed by
only $S = 51$ group-level observations. Sluggish mixing here would undermine
the entire random-effect structure. The trace plots below confirm that all
four chains explore the same region of parameter space.

```{r fig-trace-tau, eval = has_data, fig.height = 7, fig.width = 8, fig.cap = "MCMC trace plots for all six random-effect standard deviations (tau) across four chains. The extensive-margin intercept tau (top-left) shows the widest dynamic range. All panels display overlapping, stationary traces with no visible trends or stuck chains."}
if (requireNamespace("ggplot2", quietly = TRUE)) {
  library(ggplot2)

  draws <- results$draws_df
  tau_params <- paste0("tau[", 1:6, "]")
  tau_display <- c(
    "tau[1]" = "\u03c4(\u03b1 : Intercept)",
    "tau[2]" = "\u03c4(\u03b1 : Poverty)",
    "tau[3]" = "\u03c4(\u03b1 : Urban)",
    "tau[4]" = "\u03c4(\u03b2 : Intercept)",
    "tau[5]" = "\u03c4(\u03b2 : Poverty)",
    "tau[6]" = "\u03c4(\u03b2 : Urban)"
  )

  trace_data <- do.call(rbind, lapply(tau_params, function(p) {
    data.frame(
      iteration = draws$.iteration,
      chain     = factor(draws$.chain),
      value     = draws[[p]],
      parameter = tau_display[p],
      stringsAsFactors = FALSE
    )
  }))
  trace_data$parameter <- factor(trace_data$parameter,
                                 levels = unname(tau_display))

  ggplot(trace_data, aes(x = iteration, y = value, colour = chain)) +
    geom_line(alpha = 0.45, linewidth = 0.25) +
    facet_wrap(~ parameter, scales = "free_y", ncol = 2) +
    scale_colour_manual(
      values = c("1" = pal["extensive"], "2" = pal["intensive"],
                 "3" = pal["reference"], "4" = pal["dispersion"]),
      name = "Chain"
    ) +
    labs(x = "Iteration (post-warmup)", y = "Value",
         title = "MCMC Trace Plots for Random-Effect SDs") +
    theme_minimal(base_size = 12) +
    theme(legend.position = "bottom",
          panel.grid.minor = element_blank(),
          strip.text = element_text(face = "bold", size = 9))
}
```

All six trace plots show overlapping, stationary chains. The intercept $\tau$
parameters (top row) sample in a higher range with greater variability, while
the slope $\tau$ parameters (middle and bottom rows) concentrate near zero
with occasional excursions. The absence of divergent transitions (see the
diagnostic summary above) confirms that the sampler navigated the hierarchical
posterior geometry successfully.

For a complementary view, the $\hat{R}$ and ESS values in the $\tau$ table
provide formal convergence diagnostics. Slightly noisier traces for $\tau$ are
expected because these parameters are constrained positive and informed by
only $S = 51$ state-level data points.


## Model Comparison: SVC vs. Base

Does the added complexity of state-varying coefficients improve predictive
performance? We compare using PSIS-LOO:

```r
# In a live session:
loo_svc <- loo(fit_svc)
print(loo_svc)

# Compare to base model (from Vignette 01):
# comp <- hbb_loo_compare(base = fit_base, svc = fit_svc)
```

```{r load-loo, eval = has_data}
loo_obj <- readRDS(find_extdata("vig03_loo.rds"))
print(loo_obj)
```

Key quantities to examine:

- **ELPD:** Higher (less negative) means better out-of-sample prediction.
  Compare against the base model's ELPD from the *Getting Started* vignette.
- **p_loo:** The effective number of parameters --- substantially larger for
  SVC models, reflecting the 306 state-level random effects.
- **Pareto $k$:** A few observations with $k > 0.7$ are common in
  hierarchical models and usually not cause for concern.

Despite the added complexity, the SVC model should show improved or
comparable predictive performance because the hierarchical prior penalises
extreme state deviations through partial pooling, preventing overfitting.


## Complete SVC Workflow

Putting the full pipeline together:

```r
# 1. Load full dataset
data("nsece_synth", package = "hurdlebb")

# 2. Fit SVC model with survey weights
fit_svc <- hbb(
  y | trials(n_trial) ~ poverty + urban + (poverty + urban | state_id),
  data    = nsece_synth,
  weights = "weight",
  stratum = "stratum",
  psu     = "psu",
  chains  = 4,
  seed    = 42
)

# 3. Posterior summary (inspect fixed effects + tau)
summary(fit_svc)

# 4. Sandwich variance for design-corrected inference
sand <- sandwich_variance(fit_svc)
summary(fit_svc, sandwich = sand)

# 5. LOO cross-validation
loo_svc <- loo(fit_svc)

# 6. Compare to base model
comp <- hbb_loo_compare(base = fit_base, svc = fit_svc)
```


## Design Considerations

A few practical points to keep in mind when fitting SVC models:

### Sample size

The full `nsece_synth` dataset (N = 6,785 across 51 states) averages about
133 providers per state, which is adequate. The smaller `nsece_synth_small`
(N $\approx$ 500) averages only about 10 per state and is **not suitable**
for SVC models.

### Which covariates to let vary?

Not every covariate needs a random slope. If $\tau$ for a particular slope is
near zero, keep it as a fixed effect. Start with the covariates of greatest
substantive interest (here, poverty) and expand if the data support it.

### Computation time

With $P = 3$ (intercept + two covariates), the model estimates:

| Component | Count |
|:----------|------:|
| Fixed effects ($\alpha$, $\beta$, $\log\kappa$) | 7 |
| Random-effect SDs ($\tau$) | 6 |
| Cholesky correlation factors | 15 |
| State deviations ($\delta_s$) | 306 |
| **Total parameters** | **334** |

Increasing `adapt_delta` (e.g., to 0.97) and `max_treedepth` (e.g., to 14)
may be necessary if the sampler reports divergent transitions.


## Key Takeaways

1. **The SVC model reveals substantial cross-state heterogeneity.** Intercept
   $\tau$ values dominate, meaning states differ most in their baseline
   propensity to serve IT children.

2. **Slope $\tau$ values are smaller but non-negligible.** The poverty and
   urban effects are relatively consistent across states, but not perfectly
   so.

3. **Cross-margin correlations are modest.** The positive intercept
   cross-correlation suggests coherent "IT-friendly" state profiles; slope
   cross-correlations are near zero.

4. **The poverty reversal is robust.** The sign pattern persists in the global
   fixed effects, and the majority of states exhibit the reversal when
   combining the global coefficient with their state-specific deviation.

5. **MCMC convergence is adequate.** No divergent transitions, all $\hat{R}$
   close to 1, and trace plots show well-mixed chains.


## What's Next?

This vignette introduced state-varying coefficients to capture between-state
heterogeneity. The remaining vignettes extend the framework further:

| Vignette | What you will learn |
|:---------|:--------------------|
| *Policy Moderators* | Cross-level interactions: explaining *why* states differ using policy variables (CCDF market rates, tiered reimbursement, IT add-ons) |
| *Marginal Effects* | Average marginal effect decomposition across both margins, the poverty reversal probability, and design-corrected inference |
