---
title: "Policy Moderators"
author: "JoonHo Lee"
date: "`r Sys.Date()`"
description: >
  Cross-level interactions: explaining state variation in Hurdle Beta-Binomial
  coefficients using policy variables --- CCDF market rates, tiered
  reimbursement, and infant/toddler add-ons.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Policy Moderators}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
library(hurdlebb)
knitr::opts_chunk$set(collapse=TRUE, comment="#>", fig.width=7, fig.height=5, fig.align="center", out.width="85%")
set.seed(42)
pal <- c(extensive="#4393C3", intensive="#D6604D", reference="#1B7837", dispersion="#762A83")
find_extdata <- function(filename) {
  path <- system.file("extdata", filename, package = "hurdlebb")
  if (nzchar(path) && file.exists(path)) return(path)
  path2 <- file.path("..", "inst", "extdata", filename)
  if (file.exists(path2)) return(path2)
  ""
}
has_data <- nzchar(find_extdata("vig04_results.rds"))
```

## Overview

The *State-Varying Coefficients* vignette demonstrated that states differ in
how covariates relate to IT enrolment. The $\tau$ parameters quantified the
magnitude of this heterogeneity, and the caterpillar plots showed which
states deviate most from the national average. But the SVC model leaves a
fundamental question unanswered: **why do states differ?**

Policy moderators answer this by decomposing each state's deviation into a
**policy-explained component** ($\Gamma v_s$) and a **residual** ($\eta_s$).
A non-zero $\Gamma$ coefficient means that a policy variable systematically
shifts a state's regression coefficient --- connecting the SVC model to the
substantive question: do childcare subsidy policies moderate the poverty
reversal?

In this vignette you will learn how to:

1. **Understand** the three policy variables in the package and their
   substantive rationale.
2. **Formulate** the cross-level model mathematically.
3. **Specify** a policy moderator model using the `state_level()` formula
   syntax.
4. **Interpret** the $\Gamma$ coefficients --- the main results of this
   analysis.
5. **Compare** residual $\tau$ values to the pure SVC model to assess how
   much state variation is explained by policies.
6. **Evaluate** the policy model via LOO-CV.


## The Policy Variables

The package provides `nsece_state_policy`, a data frame with 51 rows (one
per state) and three policy variables drawn from the Child Care and
Development Fund (CCDF) program:

```{r load-policy-data}
data("nsece_state_policy", package = "hurdlebb")
str(nsece_state_policy)
```

```{r policy-summary}
cat("States:", nrow(nsece_state_policy), "\n")
cat("Market rate pctile (mean, SD):",
    round(mean(nsece_state_policy$mr_pctile), 2), ",",
    round(sd(nsece_state_policy$mr_pctile), 2), "\n")
cat("Tiered reimbursement:",
    round(mean(nsece_state_policy$tiered_reim), 2), "\n")
cat("IT add-on:",
    round(mean(nsece_state_policy$it_addon), 2), "\n")
```

Each variable captures a distinct dimension of state subsidy policy:

| Variable | Type | Description |
|:---------|:-----|:------------|
| `mr_pctile` | Continuous (standardised) | CCDF market rate survey percentile. Higher values mean more generous reimbursement ceilings. |
| `tiered_reim` | Binary (0/1) | Whether the state uses quality-differentiated (tiered) reimbursement rates. |
| `it_addon` | Binary (0/1) | Whether the state provides a supplemental add-on payment for infant/toddler care. |

**Market rate percentile** (`mr_pctile`). Federal guidance recommends
that states set CCDF reimbursement rates at or above the 75th percentile
of local market rates. Higher values mean providers are compensated closer
to private-market prices, reducing disincentives to accept subsidised
children. Standardised to mean 0, SD 1 for MCMC efficiency.

**Tiered reimbursement** (`tiered_reim`). States that pay higher rates to
providers meeting quality standards. This may differentially affect centres
serving IT children, who require more intensive staffing.

**IT add-on** (`it_addon`). A supplemental payment for infant/toddler
care, recognising that lower staff-to-child ratios make younger children
costlier to serve. This is the most targeted policy instrument --- it
directly addresses the cost barrier central to the poverty reversal.


## Visualising the Policy Landscape

Before fitting the model, it is useful to see how the three policy
variables are distributed across the 51 states. This context helps
interpret the $\Gamma$ coefficients later --- a variable with limited
variation across states provides less statistical leverage to detect a
moderating effect.

```{r fig-policy-dist, fig.height = 4.5, fig.width = 8.5, fig.cap = "Distribution of the three policy variables across 51 states. Left: CCDF market rate percentile (continuous, standardised). Centre: tiered reimbursement (binary, 84 percent prevalence). Right: IT add-on payment (binary, 24 percent prevalence). The heavy skew in tiered reimbursement limits statistical power for that moderator."}
if (requireNamespace("ggplot2", quietly = TRUE)) {
  library(ggplot2)

  pol <- nsece_state_policy

  # --- Panel 1: mr_pctile histogram ---
  mr_mean <- mean(pol$mr_pctile)
  p1 <- ggplot(pol, aes(x = mr_pctile)) +
    geom_histogram(bins = 13, fill = pal["extensive"], colour = "white",
                   alpha = 0.85, linewidth = 0.3) +
    geom_vline(xintercept = mr_mean, linetype = "dashed",
               colour = "grey30", linewidth = 0.5) +
    annotate("text", x = mr_mean, y = Inf,
             label = sprintf("mean = %.2f", mr_mean),
             vjust = 1.8, hjust = -0.1, size = 3, colour = "grey30") +
    labs(x = "Market rate percentile (standardised)",
         y = "Number of states",
         subtitle = "mr_pctile (continuous)") +
    theme_minimal(base_size = 10) +
    theme(panel.grid.minor = element_blank(),
          plot.subtitle = element_text(face = "bold", size = 10))

  # --- Panel 2: tiered_reim bar chart ---
  tiered_tbl <- data.frame(
    status = factor(c("No (0)", "Yes (1)"), levels = c("No (0)", "Yes (1)")),
    count  = c(sum(pol$tiered_reim == 0), sum(pol$tiered_reim == 1))
  )
  p2 <- ggplot(tiered_tbl, aes(x = status, y = count)) +
    geom_col(fill = pal["intensive"], alpha = 0.85, width = 0.55) +
    geom_text(aes(label = paste0(count, " (",
                                  round(100 * count / sum(count)), "%)")),
              vjust = -0.5, size = 3.2) +
    labs(x = NULL, y = "Number of states",
         subtitle = "tiered_reim (binary)") +
    scale_y_continuous(expand = expansion(mult = c(0, 0.18))) +
    theme_minimal(base_size = 10) +
    theme(panel.grid.minor = element_blank(),
          panel.grid.major.x = element_blank(),
          plot.subtitle = element_text(face = "bold", size = 10))

  # --- Panel 3: it_addon bar chart ---
  addon_tbl <- data.frame(
    status = factor(c("No (0)", "Yes (1)"), levels = c("No (0)", "Yes (1)")),
    count  = c(sum(pol$it_addon == 0), sum(pol$it_addon == 1))
  )
  p3 <- ggplot(addon_tbl, aes(x = status, y = count)) +
    geom_col(fill = pal["dispersion"], alpha = 0.85, width = 0.55) +
    geom_text(aes(label = paste0(count, " (",
                                  round(100 * count / sum(count)), "%)")),
              vjust = -0.5, size = 3.2) +
    labs(x = NULL, y = "Number of states",
         subtitle = "it_addon (binary)") +
    scale_y_continuous(expand = expansion(mult = c(0, 0.18))) +
    theme_minimal(base_size = 10) +
    theme(panel.grid.minor = element_blank(),
          panel.grid.major.x = element_blank(),
          plot.subtitle = element_text(face = "bold", size = 10))

  # --- Combine ---
  if (requireNamespace("patchwork", quietly = TRUE)) {
    library(patchwork)
    (p1 + p2 + p3) +
      plot_annotation(
        title = "State Policy Variable Distributions (N = 51)",
        theme = theme(plot.title = element_text(face = "bold", size = 12))
      )
  } else if (requireNamespace("gridExtra", quietly = TRUE)) {
    gridExtra::grid.arrange(p1, p2, p3, ncol = 3)
  } else {
    print(p1)
  }
}
```

Several features are worth noting:

- **Market rate percentile** spans the full standardised range with
  adequate spread, providing good statistical leverage for detecting
  moderation.
- **Tiered reimbursement** is heavily skewed (approximately 84% adoption),
  leaving few states in the "No" group. Gamma estimates for this
  moderator will be less precise.
- **IT add-on** (approximately 24% adoption) provides a more balanced
  contrast, making it the most informative binary moderator in terms of
  statistical power.


## The Cross-Level Model

In the SVC model from the previous vignette, the state random effects
followed:

$$\delta_s \sim \text{MVN}(0, \Sigma), \qquad s = 1, \ldots, S.$$

The policy moderator model replaces this with a **conditional mean**
structure:

$$\delta_s = \Gamma \, v_s + \eta_s, \qquad \eta_s \sim \text{MVN}(0, \Sigma_\eta),$$

where:

- $v_s = (1, \text{mr\_pctile}_s, \text{tiered\_reim}_s, \text{it\_addon}_s)' \in \mathbb{R}^Q$
  is the **policy vector** for state $s$. The leading 1 allows each
  random-effect dimension to have its own residual intercept.
- $\Gamma \in \mathbb{R}^{2P \times Q}$ is the **cross-level coefficient
  matrix**. Each entry $\Gamma_{k,q}$ tells us how policy variable $q$
  shifts the $k$-th random-effect dimension.
- $\eta_s$ is the **residual** state deviation --- the part of state
  heterogeneity not explained by the observed policy variables.
- $\Sigma_\eta$ is the **residual covariance** among the random effects
  after conditioning on policies.

With $P = 3$ and $Q = 4$, $\Gamma$ is a $6 \times 4$ matrix (24 elements).
Rows are random-effect dimensions; columns are policy variables.

**Example.** If $\Gamma_{2,4} > 0$ (row: $\alpha$-poverty, column:
IT add-on), then states with IT add-ons have a **less negative** poverty
effect on participation --- the add-on mitigates the poverty barrier.

The total variance decomposes as
$\text{Var}(\delta_s) = \Gamma \, \text{Var}(v_s) \, \Gamma' + \Sigma_\eta$,
separating **policy-explained** from **residual** variance.


## Formula Syntax

In **hurdlebb**, policy moderators are specified using the `state_level()`
function within the model formula, and the state-level data is passed
through the `state_data` argument:

```r
# This is the code you would run (takes ~60-90 minutes on full data):
fit_policy <- hbb(
  y | trials(n_trial) ~ poverty + urban +
    (poverty + urban | state_id) +
    state_level(mr_pctile + tiered_reim + it_addon),
  data       = nsece_synth,
  state_data = nsece_state_policy,
  weights    = "weight",
  stratum    = "stratum",
  psu        = "psu",
  chains     = 4,
  seed       = 42
)
```

Key points:

- **`state_level()` adds the $V$ matrix.** The policy names inside
  `state_level()` are looked up in `state_data`. An intercept column is
  auto-prepended, producing a $51 \times 4$ matrix $V$.
- **`state_data` is required.** Policy variables live in a separate data
  frame (one row per state), linked via `state_id`.
- **Random effects persist.** The `(poverty + urban | state_id)` term
  remains; policy moderators explain part of the random-effect variation,
  not replace it.
- **Survey weights compose cleanly** --- identical to previous vignettes.
- **Runtime.** Expect 60--90 minutes (24 extra $\Gamma$ parameters).


## Loading Pre-computed Results

As with previous vignettes, we load saved output to avoid lengthy MCMC runs:

```{r load-data}
data("nsece_synth", package = "hurdlebb")
data("nsece_state_policy", package = "hurdlebb")
```

```{r load-results, eval = has_data}
results <- readRDS(find_extdata("vig04_results.rds"))
```

The `print()` method gives a compact overview:

```{r show-print, eval = has_data}
cat(paste(results$print_text, collapse = "\n"))
```


## Model Summary

The fixed effects represent population-average coefficients after accounting
for state random effects and policy moderators:

```{r show-summary, eval = has_data}
cat(paste(results$summary_text, collapse = "\n"))
```

### Fixed effects

```{r coef-table, eval = has_data}
results$coef_both
```

The poverty reversal persists:

```{r poverty-reversal, eval = has_data}
cat(sprintf("Extensive margin (alpha_poverty): %+.3f\n",
            results$coef_both["alpha_poverty"]))
cat(sprintf("Intensive margin  (beta_poverty): %+.3f\n",
            results$coef_both["beta_poverty"]))
```

This is expected: the $\Gamma$ matrix operates on the *random effects*, not
the fixed effects. $\Gamma$ explains variation in how states deviate from
the population-average $\alpha$ and $\beta$.


## Gamma Coefficients: The Main Results

The $\Gamma$ matrix is the centrepiece of this analysis. Each entry tells
us how a state-level policy variable shifts a particular random-effect
dimension. The `gamma_table` contains all 24 entries with posterior
summaries:

```{r gamma-table, eval = has_data}
gt <- results$gamma_table
knitr::kable(
  gt[, c("parameter", "policy", "margin", "estimate", "sd",
         "ci_lower", "ci_upper", "rhat", "ess_bulk")],
  digits  = 3,
  align   = c("l", "l", "l", rep("r", 6)),
  caption = "Gamma coefficients: cross-level policy effects on state-varying parameters. 90% credible intervals (5th and 95th percentiles)."
)
```

Each row is one $\Gamma_{k,q}$ element. A credible interval excluding zero
indicates that the policy variable has a discernible effect on that
random-effect dimension.


## Gamma Heatmap

The heatmap provides a compact, publication-quality summary of all 18
cross-level interaction effects (6 parameters $\times$ 3 policy variables,
excluding the intercept column of $\Gamma$ which is a nuisance parameter).
Row grouping by margin immediately reveals whether policy moderation
operates differently on participation versus intensity.

```{r fig-gamma-heatmap, eval = has_data, fig.height = 5.5, fig.width = 7.5, fig.cap = "Heatmap of gamma coefficients (cross-level interactions). Rows are the six state-varying parameters grouped by margin (extensive on top, intensive on bottom). Columns are the three policy variables. Diverging colour scale: blue = negative, white = zero, red = positive. Signed point estimates annotated in each cell. Bold black borders mark coefficients whose 90% CI excludes zero. The heavy horizontal line separates the two margin blocks."}
if (requireNamespace("ggplot2", quietly = TRUE)) {
  library(ggplot2)

  gt <- results$gamma_table

  # Filter to non-intercept policy columns (the substantive interactions)
  # gamma_table$policy values: "intercept", "mr_pctile", "tiered_reim", "it_addon"
  gt_pol <- gt[gt$policy != "intercept", ]

  # Create display labels from the 'parameter' column
  # parameter format: "alpha_(Intercept)", "alpha_poverty", "alpha_urban",
  #                   "beta_(Intercept)", "beta_poverty", "beta_urban"
  gt_pol$re_label <- gsub("^alpha_", "\u03b1 : ", gt_pol$parameter)
  gt_pol$re_label <- gsub("^beta_",  "\u03b2 : ", gt_pol$re_label)
  gt_pol$re_label <- gsub("\\(Intercept\\)", "Intercept", gt_pol$re_label)

  # Fixed ordering: extensive (top) then intensive (bottom)
  param_order <- c(
    "\u03b1 : Intercept", "\u03b1 : poverty", "\u03b1 : urban",
    "\u03b2 : Intercept", "\u03b2 : poverty", "\u03b2 : urban"
  )
  gt_pol$re_label <- factor(gt_pol$re_label, levels = rev(param_order))

  # Readable policy axis labels
  policy_labels <- c(
    "mr_pctile"   = "Market Rate\nPercentile",
    "tiered_reim" = "Tiered\nReimbursement",
    "it_addon"    = "IT Add-On\nPayment"
  )
  gt_pol$policy_label <- factor(
    policy_labels[gt_pol$policy],
    levels = unname(policy_labels)
  )

  # Dynamic text colour: white on strongly coloured tiles
  max_abs <- max(abs(gt_pol$estimate), na.rm = TRUE)
  gt_pol$text_col <- ifelse(
    abs(gt_pol$estimate) > max_abs * 0.45,
    "white", "grey20"
  )

  # Significance flag (90% CI excludes zero)
  gt_pol$significant <- (gt_pol$ci_lower > 0) | (gt_pol$ci_upper < 0)

  ggplot(gt_pol, aes(x = policy_label, y = re_label, fill = estimate)) +
    geom_tile(colour = "white", linewidth = 0.7) +
    # Bold outline for significant coefficients
    geom_tile(data = gt_pol[gt_pol$significant, ],
              aes(x = policy_label, y = re_label),
              fill = NA, colour = "black", linewidth = 1.2) +
    # Signed point estimates
    geom_text(aes(label = sprintf("%+.3f", estimate),
                  colour = text_col),
              size = 3.2, fontface = "bold", show.legend = FALSE) +
    scale_fill_gradient2(
      low      = pal["extensive"],
      mid      = "white",
      high     = pal["intensive"],
      midpoint = 0,
      name     = expression(hat(gamma))
    ) +
    scale_colour_identity() +
    # Heavy line separating extensive / intensive blocks
    geom_hline(yintercept = 3.5, colour = "grey25", linewidth = 0.9) +
    # Margin block labels
    annotate("text", x = 3.6, y = 5, label = "Extensive",
             size = 3.2, fontface = "bold", colour = pal["extensive"],
             hjust = 0) +
    annotate("text", x = 3.6, y = 2, label = "Intensive",
             size = 3.2, fontface = "bold", colour = pal["intensive"],
             hjust = 0) +
    coord_cartesian(clip = "off") +
    labs(x = NULL, y = NULL,
         title = expression("Cross-Level Interaction Coefficients (" *
                             Gamma * " matrix)"),
         subtitle = paste("Bold borders = 90% CI excludes zero;",
                          "18 policy-specific interactions shown")) +
    theme_minimal(base_size = 12) +
    theme(panel.grid       = element_blank(),
          axis.text.x      = element_text(size = 9.5, lineheight = 1.1),
          axis.text.y      = element_text(size = 10),
          legend.position  = "right",
          plot.subtitle    = element_text(size = 9, colour = "grey45"),
          plot.margin      = margin(5, 55, 5, 5))
}
```

The heatmap reveals the pattern of policy moderation at a glance. Cells
with strong colour indicate parameter--policy pairs where the cross-level
interaction is largest. Near-zero (white) cells indicate no detectable
moderation. The separation line between the two blocks highlights whether
policy context differentially affects participation versus intensity.


## Gamma Forest Plot

For a more precise view of the gamma estimates and their uncertainty, the
forest plot shows point estimates with 90% credible intervals, faceted by
policy variable. This format makes it straightforward to identify which
interactions are credible and to compare effect magnitudes across
parameters.

```{r fig-gamma-forest, eval = has_data, fig.height = 6.5, fig.width = 8.5, fig.cap = "Forest plot of gamma coefficients by policy variable. Each panel shows how a specific policy shifts the six state-varying parameters. Filled points indicate 90% CIs that exclude zero; open points indicate CIs spanning zero. Blue = extensive margin, red = intensive margin. Dashed green line marks zero (no moderation)."}
if (requireNamespace("ggplot2", quietly = TRUE)) {
  library(ggplot2)

  gt <- results$gamma_table
  # Exclude the gamma intercept column (nuisance parameter)
  gt_pol <- gt[gt$policy != "intercept", ]

  # Create display labels from 'parameter'
  gt_pol$re_label <- gsub("^alpha_", "\u03b1 : ", gt_pol$parameter)
  gt_pol$re_label <- gsub("^beta_",  "\u03b2 : ", gt_pol$re_label)
  gt_pol$re_label <- gsub("\\(Intercept\\)", "Intercept", gt_pol$re_label)

  param_order <- c(
    "\u03b1 : Intercept", "\u03b1 : poverty", "\u03b1 : urban",
    "\u03b2 : Intercept", "\u03b2 : poverty", "\u03b2 : urban"
  )
  gt_pol$re_label <- factor(gt_pol$re_label, levels = rev(param_order))

  # Margin labels for colour
  gt_pol$Margin <- ifelse(gt_pol$margin == "extensive",
                          "Extensive", "Intensive")

  # Significance flag for shape
  gt_pol$significant <- (gt_pol$ci_lower > 0) | (gt_pol$ci_upper < 0)
  gt_pol$sig_label <- ifelse(gt_pol$significant,
                             "CI excludes zero", "CI spans zero")

  # Policy facet labels
  pol_labels <- c(mr_pctile   = "Market Rate Percentile",
                  tiered_reim = "Tiered Reimbursement",
                  it_addon    = "IT Add-On Payment")
  gt_pol$policy_facet <- factor(pol_labels[gt_pol$policy],
                                levels = unname(pol_labels))

  ggplot(gt_pol, aes(x = estimate, y = re_label, colour = Margin)) +
    geom_vline(xintercept = 0, linetype = "dashed",
               colour = pal["reference"], linewidth = 0.5) +
    # CI bars with translucent band
    geom_linerange(aes(xmin = ci_lower, xmax = ci_upper),
                   linewidth = 2, alpha = 0.25) +
    # Point estimates with significance shape encoding
    geom_point(aes(shape = sig_label), size = 2.8) +
    # Numeric annotation
    geom_text(aes(label = sprintf("%+.3f", estimate)),
              hjust = -0.25, vjust = -0.8, size = 2.5,
              show.legend = FALSE) +
    facet_wrap(~ policy_facet, scales = "free_x", ncol = 3) +
    scale_colour_manual(
      values = c(Extensive = pal["extensive"],
                 Intensive = pal["intensive"]),
      name = "Margin"
    ) +
    scale_shape_manual(
      values = c("CI excludes zero" = 16, "CI spans zero" = 1),
      name = NULL
    ) +
    labs(x = expression("Posterior estimate of " * hat(gamma)),
         y = NULL,
         title = "Policy Effects on State-Varying Coefficients",
         subtitle = "90% credible intervals (5th to 95th percentile)") +
    theme_minimal(base_size = 11) +
    theme(legend.position    = "bottom",
          panel.grid.minor   = element_blank(),
          panel.grid.major.y = element_blank(),
          strip.text         = element_text(face = "bold", size = 10),
          plot.subtitle      = element_text(size = 9, colour = "grey45"))
}
```

### Substantive interpretation

The $\Gamma$ coefficients answer specific policy questions:

- **IT add-on and the poverty slope.** $\Gamma(\alpha_\text{poverty},
  \text{it\_addon}) > 0$ would mean states with IT add-ons show a less
  negative poverty effect on participation --- targeted subsidies
  mitigating the poverty barrier.

- **Market rate and baseline participation.**
  $\Gamma(\alpha_\text{Intercept}, \text{mr\_pctile}) > 0$ indicates that
  higher reimbursement ceilings are associated with more IT service.

- **Tiered reimbursement.** Quality-differentiated rates may affect the
  two margins differently: encouraging entry versus shifting intensity.

Each non-zero $\Gamma$ represents a testable policy lever whose sign and
magnitude indicate both direction and strength of the shift.


## Tau Comparison: How Much Variance Do Policies Explain?

The clearest test of whether policy variables matter is comparing the
random-effect standard deviations ($\tau$) from the pure SVC model to the
residual $\tau$ values from the policy model. If policies explain state
variation, the residual $\tau$ should be smaller.

```{r tau-comparison-table, eval = has_data}
if (!is.null(results$tau_comparison)) {
  tc <- results$tau_comparison
  knitr::kable(
    tc[, c("parameter", "margin", "covariate",
           "tau_svc", "tau_policy", "reduction_pct")],
    digits  = 3,
    col.names = c("Parameter", "Margin", "Covariate",
                  "\u03c4 (SVC)", "\u03c4 (Policy)", "Reduction (%)"),
    align   = c("l", "l", "l", rep("r", 3)),
    caption = "Comparison of random-effect SDs between SVC and Policy models. Positive reduction means policy variables absorbed some between-state variance."
  )
}
```

```{r fig-tau-comparison, eval = has_data, fig.height = 5.5, fig.width = 7.5, fig.cap = "Comparison of random-effect standard deviations between the SVC model (open circles) and the policy moderator model (filled circles). Arrows show the direction and magnitude of change. Percentage annotations indicate the reduction in tau. Green arrows indicate reduction (policies explain variance); red arrows indicate increase."}
if (requireNamespace("ggplot2", quietly = TRUE) &&
    !is.null(results$tau_comparison)) {
  library(ggplot2)

  tc <- results$tau_comparison

  # Create parameter labels with Greek letters
  tc$label <- gsub("^alpha_", "\u03b1 : ", tc$parameter)
  tc$label <- gsub("^beta_",  "\u03b2 : ", tc$label)
  tc$label <- gsub("\\(Intercept\\)", "Intercept", tc$label)

  param_order <- c(
    "\u03b1 : Intercept", "\u03b1 : poverty", "\u03b1 : urban",
    "\u03b2 : Intercept", "\u03b2 : poverty", "\u03b2 : urban"
  )
  tc$label <- factor(tc$label, levels = rev(param_order))

  tc$Margin <- ifelse(tc$margin == "extensive", "Extensive", "Intensive")

  # Arrow colour: green for reduction, red for increase
  tc$arrow_col <- ifelse(tc$reduction_pct > 0,
                         pal["reference"], pal["intensive"])
  # Annotation text with direction arrow
  tc$pct_label <- ifelse(
    tc$reduction_pct > 0,
    sprintf("\u2193 %.1f%%", tc$reduction_pct),
    sprintf("\u2191 %.1f%%", abs(tc$reduction_pct))
  )

  ggplot(tc) +
    # Arrow from SVC tau to Policy tau
    geom_segment(aes(x = tau_svc, xend = tau_policy,
                     y = label, yend = label,
                     colour = arrow_col),
                 arrow = arrow(length = unit(0.18, "cm"), type = "closed"),
                 linewidth = 0.9, alpha = 0.7,
                 show.legend = FALSE) +
    # SVC tau (open circles)
    geom_point(aes(x = tau_svc, y = label),
               shape = 21, size = 3.5, stroke = 1,
               fill = "white", colour = "grey40") +
    # Policy tau (filled circles coloured by margin)
    geom_point(aes(x = tau_policy, y = label, fill = Margin),
               shape = 21, size = 3.2, stroke = 0.5, colour = "white") +
    # Reduction percentage annotation
    geom_text(aes(x = pmax(tau_svc, tau_policy), y = label,
                  label = pct_label, colour = arrow_col),
              hjust = -0.15, vjust = 0.4, size = 3, fontface = "bold",
              show.legend = FALSE) +
    scale_colour_identity() +
    scale_fill_manual(
      values = c(Extensive = pal["extensive"],
                 Intensive = pal["intensive"]),
      name = "Margin"
    ) +
    scale_x_continuous(expand = expansion(mult = c(0.02, 0.22))) +
    labs(x = expression(tau ~ "(random-effect SD)"),
         y = NULL,
         title = expression("Variance Reduction: SVC" %->% "Policy Model"),
         subtitle = paste("Open circle = SVC model,",
                          "filled circle = policy model")) +
    theme_minimal(base_size = 12) +
    theme(legend.position    = "bottom",
          panel.grid.minor   = element_blank(),
          panel.grid.major.y = element_blank(),
          plot.subtitle      = element_text(size = 9, colour = "grey45"))
}
```

The reduction is:
$$\text{Reduction}_k = 1 - \tau_{k,\text{policy}} / \tau_{k,\text{SVC}}.$$
A 20% reduction in $\tau$ corresponds to roughly 36% of variance explained.
Even modest reductions are substantively meaningful --- they point to
*observable* policy choices that systematically shift state coefficients.

### What the comparison reveals

- **Parameters with large reductions** indicate that the policy variables
  successfully explain part of the between-state heterogeneity. If the
  intercept $\tau$ drops substantially, differences in reimbursement
  generosity, tiered systems, or IT add-ons account for some of the
  baseline variation in IT participation or intensity.

- **Parameters with small reductions** suggest that policy context does not
  account for why states differ on that dimension. The unexplained
  heterogeneity may stem from unmeasured factors such as labour market
  conditions, regulatory burden, or cultural norms.

- **Slope reductions** are typically smaller than intercept reductions.
  This is expected: the slope $\tau$ values were already modest in the
  SVC model, leaving less variance to explain.


## Residual Tau Estimates

The residual $\tau$ values represent state heterogeneity *after* accounting
for the three policy variables:

```{r tau-table, eval = has_data}
tau <- results$tau_table
knitr::kable(
  tau[, c("parameter", "margin", "covariate", "estimate", "sd",
          "ci_lower", "ci_upper", "rhat", "ess_bulk")],
  digits  = 3,
  align   = c("l", "l", "l", rep("r", 6)),
  caption = "Residual random-effect standard deviations (policy model)."
)
```

Non-trivial residual $\tau$ values indicate that unmeasured state
characteristics (labour markets, regulations, culture) also contribute to
state heterogeneity beyond the three observed policy variables.


## LOO Comparison

We compare the policy moderator model to the SVC model using PSIS-LOO:

```r
# In a live session:
loo_policy <- loo(fit_policy)
print(loo_policy)
```

```{r load-loo, eval = has_data}
loo_obj <- readRDS(find_extdata("vig04_loo.rds"))
print(loo_obj)
```

Key quantities:

- **ELPD:** Higher (less negative) means better out-of-sample prediction.
  Compare against the SVC model's ELPD from the previous vignette.
- **p_loo:** The effective number of parameters. Adding 24 $\Gamma$
  coefficients increases complexity, but the policy-explained variance
  reduces the effective complexity of the random effects.
- **Pareto $k$:** A few high-$k$ observations are common in hierarchical
  models.

The LOO difference between SVC and policy models is typically modest
because policy variables explain *between-state* variation ($S = 51$
groups), whereas LOO evaluates observation-level prediction ($N \approx
6{,}800$). The policy model's value is **interpretive** --- explaining
*why* states differ --- not raw predictive improvement.


## Complete Workflow

```r
data("nsece_synth", package = "hurdlebb")
data("nsece_state_policy", package = "hurdlebb")

# 1. Fit
fit_policy <- hbb(
  y | trials(n_trial) ~ poverty + urban +
    (poverty + urban | state_id) +
    state_level(mr_pctile + tiered_reim + it_addon),
  data = nsece_synth, state_data = nsece_state_policy,
  weights = "weight", stratum = "stratum", psu = "psu",
  chains = 4, seed = 42
)

# 2. Inspect
summary(fit_policy)                              # fixed + Gamma
sand <- sandwich_variance(fit_policy)
summary(fit_policy, sandwich = sand)             # Wald CIs

# 3. Compare
loo_policy <- loo(fit_policy)
comp <- hbb_loo_compare(svc = fit_svc, policy = fit_policy)
```


## Design Considerations

### Choosing policy variables

Not every state-level variable belongs in the model. The policy variables
should be:

1. **Substantively motivated.** Each variable should have a plausible
   theoretical mechanism for moderating the covariate--outcome
   relationship.
2. **Variable across states.** If nearly all states have the same value
   (e.g., `tiered_reim` at 84%), there is little variation to exploit,
   and the corresponding gamma estimates will be imprecise.
3. **Measured at the state level.** Provider-level covariates enter as
   fixed effects, not as policy moderators.

### Number of policy variables

With $S = 51$ states, the model can comfortably support a handful of
state-level predictors ($Q - 1 \leq 5$ or so). Adding too many policy
variables risks overfitting the between-state component and may cause
convergence difficulties. Start with the most substantively important
variables and expand incrementally.


## Key Takeaways

1. **Policy moderators explain the "why" behind state heterogeneity.**
   The SVC model shows that states differ; the policy model identifies
   specific policy levers that drive those differences. This moves the
   analysis from description to explanation.

2. **The $\Gamma$ matrix is the main result.** Each non-zero entry
   represents a policy-to-coefficient pathway. Rows correspond to the six
   state-varying dimensions (intercept and slope on both margins); columns
   correspond to the policy variables.

3. **Tau reduction quantifies explanatory power.** Comparing residual
   $\tau$ values to the SVC model's $\tau$ values reveals how much state
   heterogeneity is absorbed by the observed policies. Even partial
   reductions are meaningful because they point to actionable policy
   variables.

4. **The poverty reversal persists at the national level.** Adding policy
   moderators does not alter the global fixed effects. The population-average
   poverty reversal remains robust.

5. **LOO improvements are modest by design.** The policy model's value is
   interpretive, not predictive. It does not dramatically change
   observation-level predictions because it operates at the state level.

6. **Residual heterogeneity remains.** The three policy variables do not
   fully explain state variation. Unmeasured state characteristics
   contribute to the residual $\tau$ values, suggesting opportunities for
   future research with additional policy indicators.


## What's Next?

This vignette introduced cross-level policy moderators to explain state-level
heterogeneity. The final vignette completes the analysis pipeline:

| Vignette | What you will learn |
|:---------|:--------------------|
| *Marginal Effects* | Average marginal effect decomposition across both margins, the poverty reversal probability, and design-corrected inference |
