---
title: "Marginal Effects"
author: "JoonHo Lee"
date: "`r Sys.Date()`"
description: >
  Average marginal effect decomposition for the Hurdle Beta-Binomial:
  separating extensive-margin (participation) and intensive-margin
  (enrolment share) contributions, with the poverty reversal probability.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Marginal Effects}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
library(hurdlebb)
knitr::opts_chunk$set(collapse=TRUE, comment="#>", fig.width=7, fig.height=5, fig.align="center", out.width="85%")
set.seed(42)
pal <- c(extensive="#4393C3", intensive="#D6604D", reference="#1B7837", dispersion="#762A83")
find_extdata <- function(filename) {
  path <- system.file("extdata", filename, package = "hurdlebb")
  if (nzchar(path) && file.exists(path)) return(path)
  path2 <- file.path("..", "inst", "extdata", filename)
  if (file.exists(path2)) return(path2)
  ""
}
has_data <- nzchar(find_extdata("vig05_results.rds"))
```

## Overview

The previous vignettes fitted Hurdle Beta-Binomial models and inspected
raw coefficients. But logit-scale coefficients are hard to interpret in
practical terms: what does an $\alpha_{\text{poverty}} = -0.14$ actually
*mean* for the expected IT enrolment share? And how much of that effect
flows through the participation decision versus the intensity decision?

**Average marginal effects (AMEs)** translate logit-scale parameters
into probability-scale changes, answering the question: *if we increase
covariate $x_k$ by one unit for every centre in the sample, by how
much does the expected IT share change on average?*

The hurdle model's two-part structure adds a crucial twist. Because the
expected IT share is the *product* of two probabilities ---
participation $q_i$ and conditional intensity $\mu_i$ --- a single
covariate can push the outcome in opposite directions across the two
margins. The AME decomposition makes this transparent by separating each
covariate's effect into its **extensive-margin** and
**intensive-margin** contributions.

In this vignette you will learn how to:

1. **Derive** the AME decomposition from the product rule.
2. **Compute** extensive, intensive, and total AMEs using `ame()`.
3. **Read** the decomposition table with sign patterns and share
   percentages.
4. **Visualise** the poverty reversal through posterior density plots.
5. **Quantify** the reversal probability from posterior draws.
6. **Optionally** obtain design-corrected Wald AME intervals via the
   sandwich variance.


## The Decomposition Formula

The Hurdle Beta-Binomial model implies that the expected IT share for
centre $i$ is:

$$E\!\left[\frac{y_i}{n_i}\right] = q_i \cdot \mu_i,$$

where $q_i = \text{logistic}(X_i'\alpha)$ is the participation
probability (extensive margin) and
$\mu_i = \text{logistic}(X_i'\beta)$ is the conditional enrolment share
(intensive margin).

Differentiating with respect to covariate $x_{ik}$ using the product
rule:

$$\frac{\partial\, E[y_i / n_i]}{\partial\, x_{ik}}
= \underbrace{\alpha_k \, q_i(1 - q_i) \, \mu_i}_{\text{extensive-margin AME}}
\;+\; \underbrace{\beta_k \, \mu_i(1 - \mu_i) \, q_i}_{\text{intensive-margin AME}}.$$

The first term captures how much $x_k$ changes IT share by altering
*whether* the centre participates. The second captures how much $x_k$
changes IT share by altering *how intensely* participating centres
serve IT children.

The **average marginal effect** averages over all $N$ observations:

$$\text{AME}_k = \frac{1}{N} \sum_{i=1}^{N}
\bigl[\text{ext}_{ik} + \text{int}_{ik}\bigr]
= \text{AME}_k^{\text{ext}} + \text{AME}_k^{\text{int}}.$$

A key computational insight: the $N$-vectors $q_i(1-q_i)\mu_i$ and
$\mu_i(1-\mu_i)q_i$ do not depend on which covariate $k$ we are
differentiating with respect to. So we compute them once per MCMC draw
and multiply by the appropriate coefficient:

$$\text{AME}_k^{\text{ext}} = \alpha_k \cdot \overline{q'(1-q)\,\mu},
\qquad
\text{AME}_k^{\text{int}} = \beta_k \cdot \overline{\mu'(1-\mu)\,q},$$

reducing the per-draw cost from $O(NP)$ to $O(N)$.


## Computing AMEs

The `ame()` function handles the full computation --- looping over
posterior draws, computing extensive and intensive kernels, and
returning point estimates, credible intervals, and reversal
probabilities.

```r
# In a live session after fitting:
fit <- hbb(y | trials(n_trial) ~ poverty + urban,
           data = nsece_synth, weights = "weight",
           stratum = "stratum", psu = "psu",
           chains = 4, seed = 42)

# Basic AME decomposition
ame_result <- ame(fit)

# With design correction (adds Wald CIs):
# ame_result <- ame(fit, cholesky = chol, sandwich = sand)
```

We load pre-computed results for this vignette:

```{r load-results, eval = has_data}
results <- readRDS(find_extdata("vig05_results.rds"))
ame_obj <- readRDS(find_extdata("vig05_ame.rds"))
```


## AME Print Output

The `print()` method provides a structured summary of the decomposition:

```{r show-print, eval = has_data}
cat(paste(results$ame_print_text, collapse = "\n"))
```

The output reports three levels: **population-level predictions**
($\bar{q} \approx 0.60$, $\bar{\mu} \approx 0.49$,
product $\approx 0.29$), the **decomposition table** with extensive
and intensive components and sign patterns, and **reversal
probabilities** for each non-intercept covariate.


## Decomposition Table

The decomposition table is the centrepiece of the analysis. Extract it
with `ame_decomposition()`:

```{r decomp-table, eval = has_data}
decomp <- results$decomp_df
knitr::kable(
  decomp[, c("covariate", "ext_ame", "int_ame", "total_ame",
             "ext_share", "int_share", "sign_pattern")],
  digits  = 4,
  col.names = c("Covariate", "Ext AME", "Int AME", "Total AME",
                "Ext %", "Int %", "Pattern"),
  align   = c("l", "r", "r", "r", "r", "r", "l"),
  caption = "AME decomposition: extensive and intensive contributions."
)
```

### Reading the table

Each row shows how a one-unit increase in the covariate changes the
expected IT share, decomposed into:

- **Ext AME:** The contribution through participation probability
  ($\alpha_k \cdot \overline{q'(1-q)\mu}$).
- **Int AME:** The contribution through conditional intensity
  ($\beta_k \cdot \overline{\mu'(1-\mu)q}$).
- **Total AME:** The sum of the two components.
- **Ext / Int %:** Each component's share of the total absolute
  magnitude.
- **Pattern:** Whether the two components push in the same direction
  ("reinforcing") or opposite directions ("opposing").

Both covariates display an **opposing** pattern, meaning their
extensive and intensive effects work against each other. This is the
hallmark of a hurdle model where ignoring the two-part structure would
produce a misleading single coefficient.


## The Poverty Reversal

The poverty covariate exemplifies the key insight of the AME
decomposition. Looking at the components:

```{r poverty-detail, eval = has_data}
pov <- decomp[decomp$covariate == "poverty", ]
cat(sprintf("Poverty extensive AME:  %+.4f\n", pov$ext_ame))
cat(sprintf("Poverty intensive AME:  %+.4f\n", pov$int_ame))
cat(sprintf("Poverty total AME:      %+.4f\n", pov$total_ame))
cat(sprintf("Extensive share:        %.1f%%\n", pov$ext_share))
cat(sprintf("Intensive share:        %.1f%%\n", pov$int_share))
cat(sprintf("Sign pattern:           %s\n",     pov$sign_pattern))
```

The extensive-margin AME is **negative**: a one-unit increase in
neighbourhood poverty *reduces* the probability that a centre serves IT
children. But the intensive-margin AME is **positive**: among centres
that do serve, the same increase in poverty *raises* the IT share.

This is the **poverty reversal**. The total AME is the net of these two
opposing forces. The extensive margin dominates (about 59% of the total
absolute magnitude), so the net effect is a small negative --- poverty
slightly reduces overall IT shares, but the reduction comes entirely
from the participation decision, not from intensity.

A standard single-equation model would report only the total, hiding
the reversal entirely. The decomposition reveals that poverty has two
qualitatively different effects operating through distinct mechanisms.


## Posterior Density: Poverty AME

The posterior draws let us visualise the full uncertainty around each
component. The density plot below overlays the extensive, intensive, and
total AME distributions for the poverty covariate:

```{r fig-poverty-density, eval = has_data, fig.height = 5, fig.cap = "Posterior densities of the poverty AME decomposition. Blue = extensive margin, red = intensive margin, green = total. Vertical dashed lines mark posterior means; the solid grey line marks zero."}
if (requireNamespace("ggplot2", quietly = TRUE)) {
  library(ggplot2)
  dens_df <- rbind(
    data.frame(value = results$poverty_ext_draws,   component = "Extensive"),
    data.frame(value = results$poverty_int_draws,   component = "Intensive"),
    data.frame(value = results$poverty_total_draws, component = "Total"))
  dens_df$component <- factor(dens_df$component,
                               levels = c("Extensive", "Intensive", "Total"))
  comp_colours <- c(Extensive = pal["extensive"], Intensive = pal["intensive"],
                    Total = pal["reference"])
  mu_ext <- mean(results$poverty_ext_draws)
  mu_int <- mean(results$poverty_int_draws)
  means <- data.frame(
    component = factor(c("Extensive", "Intensive", "Total"),
                       levels = c("Extensive", "Intensive", "Total")),
    xval = c(mu_ext, mu_int, mean(results$poverty_total_draws)))

  ggplot(dens_df, aes(x = value, fill = component, colour = component)) +
    geom_density(alpha = 0.25, linewidth = 0.8) +
    geom_vline(xintercept = 0, colour = "grey40", linewidth = 0.7) +
    geom_vline(data = means, aes(xintercept = xval, colour = component),
               linetype = "dashed", linewidth = 0.6, show.legend = FALSE) +
    scale_fill_manual(values = comp_colours, name = "Component") +
    scale_colour_manual(values = comp_colours, name = "Component") +
    annotate("text", x = mu_ext - 0.015, y = Inf, vjust = 1.5,
             label = sprintf("%.4f", mu_ext),
             colour = pal["extensive"], size = 3, fontface = "italic") +
    annotate("text", x = mu_int + 0.015, y = Inf, vjust = 1.5,
             label = sprintf("%.4f", mu_int),
             colour = pal["intensive"], size = 3, fontface = "italic") +
    labs(x = "AME (probability-scale change per unit)",
         y = "Posterior density",
         title = "Poverty AME Decomposition: Posterior Densities") +
    theme_minimal(base_size = 12) +
    theme(legend.position = "bottom", panel.grid.minor = element_blank())
}
```

The extensive (blue) and intensive (red) densities straddle zero on
opposite sides, visualising the opposing-sign pattern. The total (green)
distribution is wider --- it inherits variability from both margins ---
and centres near zero, reflecting near-cancellation of opposing forces.


## Reversal Probability

The **reversal probability** formalises the sign pattern --- the
fraction of posterior draws where the extensive and intensive AMEs
have opposing signs:
$$\Pr(\text{reversal}_k) = \Pr\!\bigl(\text{AME}_k^{\text{ext}} < 0
\;\text{AND}\; \text{AME}_k^{\text{int}} > 0\bigr)
+ \Pr\!\bigl(\text{AME}_k^{\text{ext}} > 0
\;\text{AND}\; \text{AME}_k^{\text{int}} < 0\bigr).$$

```{r reversal-probs, eval = has_data}
cat("Reversal probabilities:\n")
cat(sprintf("  poverty: %.3f\n", ame_obj$reversal_probs["poverty"]))
cat(sprintf("  urban:   %.3f\n", ame_obj$reversal_probs["urban"]))
```

```{r fig-reversal, eval = has_data, fig.height = 4.5, fig.cap = "Scatter of posterior draws for the poverty AME. Shaded quadrants are reversal regions (opposing signs). Their fraction is the reversal probability."}
if (requireNamespace("ggplot2", quietly = TRUE)) {
  library(ggplot2)
  n_show <- min(2000, length(results$poverty_ext_draws))
  scatter_df <- data.frame(
    ext = results$poverty_ext_draws[seq_len(n_show)],
    int = results$poverty_int_draws[seq_len(n_show)])
  scatter_df$quad <- ifelse(
    (scatter_df$ext < 0 & scatter_df$int > 0) |
    (scatter_df$ext > 0 & scatter_df$int < 0),
    "Opposing (reversal)", "Same sign")
  rev_lab <- sprintf("Pr(reversal) = %.3f", ame_obj$reversal_probs["poverty"])

  ggplot(scatter_df, aes(x = ext, y = int, colour = quad)) +
    annotate("rect", xmin = -Inf, xmax = 0, ymin = 0, ymax = Inf,
             fill = pal["dispersion"], alpha = 0.06) +
    annotate("rect", xmin = 0, xmax = Inf, ymin = -Inf, ymax = 0,
             fill = pal["dispersion"], alpha = 0.06) +
    geom_hline(yintercept = 0, colour = "grey50", linewidth = 0.5) +
    geom_vline(xintercept = 0, colour = "grey50", linewidth = 0.5) +
    geom_point(alpha = 0.2, size = 0.8) +
    scale_colour_manual(values = c("Opposing (reversal)" = pal["dispersion"],
                                   "Same sign" = "grey55"), name = NULL) +
    annotate("text", x = min(scatter_df$ext) * 0.6,
             y = max(scatter_df$int) * 0.85, label = rev_lab,
             size = 4.5, fontface = "bold", colour = pal["dispersion"]) +
    labs(x = "Extensive-margin AME (poverty)",
         y = "Intensive-margin AME (poverty)",
         title = "Poverty Reversal: Posterior Draw Scatter") +
    theme_minimal(base_size = 12) +
    theme(legend.position = "bottom", panel.grid.minor = element_blank())
}
```

The shaded quadrants mark reversal regions (opposing signs). In the
synthetic data about half the posterior mass falls in these regions. In
the full NSECE analysis the reversal probability approaches 1.000,
reflecting the much stronger signal in the real data.


## Urban AME

The urban covariate also shows an opposing pattern, though the
decomposition is more one-sided:

```{r urban-detail, eval = has_data}
urb <- decomp[decomp$covariate == "urban", ]
cat(sprintf("Urban extensive AME:  %+.4f\n", urb$ext_ame))
cat(sprintf("Urban intensive AME:  %+.4f\n", urb$int_ame))
cat(sprintf("Urban total AME:      %+.4f\n", urb$total_ame))
cat(sprintf("Extensive share:      %.1f%%\n", urb$ext_share))
```

Urban location raises participation probability (positive extensive
AME) but slightly reduces conditional IT intensity (negative intensive
AME). The extensive margin dominates overwhelmingly (about 91% of the
total magnitude), so the net urban effect is positive. The intensive
component is so small that the total AME is essentially driven by the
participation channel alone.


## AME Decomposition Bar Chart

A stacked bar chart provides a compact visual summary of how each
covariate's effect decomposes across the two margins:

```{r fig-bar-decomp, eval = has_data, fig.height = 4.5, fig.cap = "AME decomposition bar chart. Extensive (blue) and intensive (red) contributions for each covariate; green diamonds mark the total (net) AME."}
if (requireNamespace("ggplot2", quietly = TRUE)) {
  library(ggplot2)
  bar_df <- rbind(
    data.frame(covariate = decomp$covariate, component = "Extensive",
               value = decomp$ext_ame),
    data.frame(covariate = decomp$covariate, component = "Intensive",
               value = decomp$int_ame))
  bar_df$component <- factor(bar_df$component,
                              levels = c("Extensive", "Intensive"))
  total_df <- data.frame(covariate = decomp$covariate,
                          total = decomp$total_ame)

  ggplot(bar_df, aes(x = covariate, y = value, fill = component)) +
    geom_col(position = position_dodge(width = 0.7),
             width = 0.6, alpha = 0.85) +
    geom_point(data = total_df, aes(x = covariate, y = total),
               inherit.aes = FALSE, shape = 18, size = 4,
               colour = pal["reference"]) +
    geom_hline(yintercept = 0, colour = "grey40", linewidth = 0.6) +
    scale_fill_manual(values = c(Extensive = pal["extensive"],
                                  Intensive = pal["intensive"]),
                       name = "Component") +
    labs(x = NULL, y = "Average Marginal Effect",
         title = "AME Decomposition by Covariate",
         subtitle = "Green diamonds = total (net) AME") +
    theme_minimal(base_size = 12) +
    theme(legend.position = "bottom", panel.grid.minor = element_blank(),
          plot.subtitle = element_text(size = 9, colour = "grey45"))
}
```

Poverty's bars point in opposite directions --- the visual signature of
the reversal. Urban's decomposition is heavily asymmetric: the
extensive component dominates with only a tiny offsetting intensive
component.


## AME Forest Plot

A forest plot displays point estimates with credible intervals across
all margins and covariates simultaneously:

```{r fig-forest-ame, eval = has_data, fig.height = 5, fig.cap = "Forest plot of AME estimates with 95% credible intervals. Extensive (blue), intensive (red), and total (green) components for each covariate."}
if (requireNamespace("ggplot2", quietly = TRUE)) {
  library(ggplot2)

  make_forest_row <- function(summ, comp_name) {
    s <- summ[summ$covariate != "(Intercept)", ]
    data.frame(covariate = s$covariate, component = comp_name,
               estimate = s$post_mean, ci_lo = s$ci_lo, ci_hi = s$ci_hi)
  }
  forest_df <- rbind(
    make_forest_row(ame_obj$ext_summary,   "Extensive"),
    make_forest_row(ame_obj$int_summary,   "Intensive"),
    make_forest_row(ame_obj$total_summary, "Total")
  )
  forest_df$component <- factor(forest_df$component,
                                 levels = c("Extensive", "Intensive", "Total"))
  forest_df$covariate <- factor(forest_df$covariate,
                                 levels = rev(unique(forest_df$covariate)))
  forest_df$ypos <- as.numeric(forest_df$covariate) +
    c(Extensive = 0.2, Intensive = 0, Total = -0.2)[as.character(forest_df$component)]

  comp_colours <- c(Extensive = pal["extensive"],
                    Intensive = pal["intensive"],
                    Total     = pal["reference"])

  ggplot(forest_df, aes(x = estimate, y = ypos, colour = component)) +
    geom_vline(xintercept = 0, linetype = "dashed", colour = "grey50",
               linewidth = 0.6) +
    geom_pointrange(aes(xmin = ci_lo, xmax = ci_hi),
                    size = 0.5, linewidth = 0.7, fatten = 2.5) +
    geom_text(aes(label = sprintf("%+.4f", estimate)),
              hjust = -0.3, vjust = -0.6, size = 2.8,
              show.legend = FALSE) +
    scale_colour_manual(values = comp_colours, name = "Component") +
    scale_y_continuous(breaks = seq_along(levels(forest_df$covariate)),
                       labels = levels(forest_df$covariate)) +
    labs(x = "Average Marginal Effect", y = NULL,
         title = "AME Forest Plot with 95% Credible Intervals") +
    theme_minimal(base_size = 12) +
    theme(legend.position = "bottom", panel.grid.minor = element_blank(),
          panel.grid.major.y = element_blank())
}
```

The wide credible intervals reflect posterior uncertainty in the
synthetic dataset. In the full NSECE analysis (N = 6,785), these
intervals are considerably narrower and the poverty reversal emerges
with much greater precision.


## Population-Level Predictions

The AME object stores per-draw mean participation probability $\bar{q}$
and mean conditional intensity $\bar{\mu}$, providing context for
interpreting AME magnitudes:

```{r pop-predictions, eval = has_data}
cat(sprintf("Mean P(serve IT) = %.4f,  Mean E[IT share | serve] = %.4f\n",
            mean(ame_obj$mean_q), mean(ame_obj$mean_mu)))
cat(sprintf("Mean E[IT share] = %.4f\n",
            mean(ame_obj$mean_q) * mean(ame_obj$mean_mu)))
```

A poverty extensive AME of $-0.010$ against a baseline $\bar{q}
\approx 0.60$ means poverty reduces participation by about 1.7
percentage points per unit on the probability scale.


## Design-Corrected AME (Optional)

When working with complex survey data, posterior-based AME credible
intervals may understate uncertainty (see *Survey Design* vignette).
Pass `cholesky` and `sandwich` arguments to `ame()` for
design-corrected Wald CIs via the delta method:

```r
sand <- sandwich_variance(fit)
chol <- cholesky_correct(fit, sand)
ame_corrected <- ame(fit, cholesky = chol, sandwich = sand)
```

The Wald variance is
$\text{Var}(\text{AME}_k) \approx
\nabla_\theta \text{AME}_k(\hat\theta)^\top
V_{\text{sand}}
\nabla_\theta \text{AME}_k(\hat\theta)$,
where the gradient is computed by central differences ($\epsilon =
10^{-5}$). The print output includes a "Wald AME" section with a
sign-agreement check between posterior and Wald approaches.


## Complete Workflow

```r
# 1. Fit the model
fit <- hbb(y | trials(n_trial) ~ poverty + urban,
           data = nsece_synth, weights = "weight",
           stratum = "stratum", psu = "psu", chains = 4, seed = 42)

# 2. (Optional) Sandwich + Cholesky correction
sand <- sandwich_variance(fit)
chol <- cholesky_correct(fit, sand)

# 3. AME decomposition (with or without design correction)
ame_result <- ame(fit, cholesky = chol, sandwich = sand)
print(ame_result)

# 4. Extract decomposition table
ame_decomposition(ame_result)

# 5. Reversal probabilities
ame_result$reversal_probs

# 6. Posterior draws for custom analysis (M_use x P matrices)
str(ame_result$ext_ame_draws)
str(ame_result$int_ame_draws)
```


## Key Takeaways

1. **The AME decomposition separates extensive and intensive channels.**
   A single covariate effect is split into a participation component
   ($\alpha_k \cdot \overline{q'(1-q)\mu}$) and an intensity component
   ($\beta_k \cdot \overline{\mu'(1-\mu)q}$). This decomposition is
   unique to the hurdle model and impossible to recover from a
   single-equation specification.

2. **Opposing signs reveal hidden heterogeneity.** When the two
   components push in opposite directions (the "opposing" or "reversal"
   pattern), a standard model would report a net effect that masks the
   true complexity. The AME decomposition makes both channels visible.

3. **The reversal probability quantifies posterior certainty.** Rather
   than relying on point-estimate sign comparisons, the reversal
   probability uses the full posterior distribution to assess how
   confidently we can claim the opposing-sign pattern holds.

4. **Design correction extends to AMEs.** The delta-method Wald
   intervals, computed via the sandwich variance, ensure that AME
   inference accounts for the survey design. This is especially
   important when the design effect ratios are large.

5. **Magnitude context matters.** Always interpret AMEs against the
   baseline population predictions ($\bar{q}$, $\bar{\mu}$). A
   seemingly small AME can represent a large proportional change
   relative to the baseline.


## Series Summary

This vignette concludes the **hurdlebb** vignette series. Across the
five vignettes, you have learned the complete analytical pipeline:

| Vignette | Topic | Key concept |
|:---------|:------|:------------|
| *Getting Started* | Core workflow | Two-part model, PPC, LOO |
| *Survey Design* | Complex surveys | Pseudo-posterior, sandwich SE, Wald CIs |
| *State-Varying Coefficients* | Hierarchical model | Random slopes, cross-margin correlations |
| *Policy Moderators* | Cross-level interactions | Why states differ in IT enrolment patterns |
| *Marginal Effects* | AME decomposition | Extensive/intensive contributions, reversal probability |

The Hurdle Beta-Binomial framework, combined with proper survey design
corrections and AME decomposition, provides a principled approach to
analysing bounded count data with excess zeros --- revealing the
distinct mechanisms through which covariates shape both participation
and intensity decisions.
